---
output:
  md_document:
    variant: markdown_github
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# brms

The <b>brms</b> package provides an interface to fit bayesian generalized linear mixed models using Stan, which is a C++ package for obtaining Bayesian inference using the No-U-turn sampler (see http://mc-stan.org/). The formula syntax is very similar to that of the package lme4 to provide a familiar and simple interface for performing regression analyses.

<!--
```{r set-options, echo=FALSE}
options(width=90)
```
-->

# How to use brms
```{r load, message=FALSE}
library(brms)
```

As a simple example, we use poisson regression to model the seizure counts in epileptic patients to investigate whether the treatment (represented by variable Trt_c) can reduce the seizure counts. Three random intercepts are incorporated to account for the variance between patients and visits, as well as for the residual variance. 
```{r fit, results='hide'}
fit <- brm(count ~ log_Age_c + log_Base4_c * Trt_c + (1|patient) + (1|visit) + (1|obs), 
           data = epilepsy, family = "poisson")
```

The results (i.e. posterior samples) can be investigated using
```{r print}
summary(fit) 
```

On the top of the output, some general information on the model is given, such as family, formula, number of iterations and chains, as well as the WAIC, which is an information criterion for Bayesian models. Next, random effects are displayed seperately for each grouping factor in terms of standard deviations and (in case of more than one random effect per grouping factor; not displayed here) correlations between random effects. On the bottom of the output, fixed effects are displayed. If incorporated, autocorrelation effects and family specific parameters (e.g., the residual standard deviation 'sigma' in normal models) are also given.

In general, every parameter is summarized using the mean ('Estimate') and the standard deviation ('Est.Error') of the posterior distribution as well as two-sided 95% Credible intervals ('l-95% CI' and 'u-95% CI') based on quantiles. The last two values ('Eff.Sample' and 'Rhat') provide information on how well the algorithm could estimate the posterior distribution of this parameter. If 'Rhat' is considerably greater than 1, the algorithm has not yet converged and it is necessary to run more iterations and / or set stronger priors.

To visually investigate the chains as well as the posterior distributions, you can use
```{r plot, eval=FALSE, fig.align='center', fig.height=8, fig.width=10}
plot(fit) 
``` 
An even more detailed investigation can be achieved by applying the shinystan package:
```{r shiny, eval=FALSE}
launch_shiny(fit) 
``` 

For a complete list of methods to apply on <b>brms</b> models see
```{r methods.brmsfit}
methods(class = "brmsfit") 
``` 

For details on formula syntax, families and link functions, as well as prior distributions see the help page of the brm function:
```{r help.brm, eval=FALSE}
help(brm) 
``` 

# How to install brms

To install the latest release version from CRAN use

```{r install_brms, eval=FALSE}
install.packages("brms")
```

The current developmental version can be downloaded from github via

```{r install_brms2, eval=FALSE}
library(devtools)
install_github("paul-buerkner/brms")
```

Because <b>brms</b> is based on Stan, a C++ compiler is required. The program Rtools (available on https://cran.r-project.org/bin/windows/Rtools/) comes with a C++ compiler for Windows. On Mac, you should use Xcode. We recommend to install rstan (available on CRAN) before installing <b>brms</b>. For further instructions see https://github.com/stan-dev/rstan/wiki/RStan-Getting-Started.