---
title: "Estimating Multivariate Models with brms"
author: "Paul Bürkner"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: yes
vignette: >
  %\VignetteIndexEntry{Estimating Multivariate Models with brms}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
params:
  EVAL: !r identical(Sys.getenv("NOT_CRAN"), "true")
---

```{r, SETTINGS-knitr, include=FALSE}
stopifnot(require(knitr))
options(width = 90)
knit_hooks$set(pngquant = knitr::hook_pngquant)
opts_chunk$set(
  comment = NA,
  message = FALSE,
  warning = FALSE,
  eval = if (isTRUE(exists("params"))) params$EVAL else FALSE,
  dev = "ragg_png",
  dpi = 72,
  fig.retina = 1.5,
  fig.asp = 0.8,
  fig.width = 5,
  out.width = "60%",
  fig.align = "center",
  pngquant = "--speed=1 --quality=50"
)
library(brms)
ggplot2::theme_set(theme_default())
```

## Introduction

In the present vignette, we want to discuss how to specify multivariate multilevel models using **brms**. We call a model *multivariate* if it contains multiple response variables, each being predicted by its own set of predictors. Consider an example from biology. Hadfield, Nutall, Osorio, and Owens (2007) analyzed data of the Eurasian blue tit (https://en.wikipedia.org/wiki/Eurasian_blue_tit). They predicted the `tarsus` length as well as the `back` color of chicks. Half of the brood were put into another `fosternest`, while the other half stayed in the fosternest of their own `dam`. This allows to separate genetic from environmental factors. Additionally, we have information about the `hatchdate` and `sex` of the chicks (the latter being known for 94\% of the animals).

```{r data}
data("BTdata", package = "MCMCglmm")
head(BTdata)
```

## Basic Multivariate Models

We begin with a relatively simple multivariate normal model.

```{r fit1, message=FALSE, warning=FALSE, results='hide'}
bform1 <-
  bf(mvbind(tarsus, back) ~ sex + hatchdate + (1 | p | fosternest) + (1 | q | dam)) +
  set_rescor(TRUE)

fit1 <- brm(bform1, data = BTdata, chains = 2, cores = 2)
```

As can be seen in the model code, we have used `mvbind` notation to tell
**brms** that both `tarsus` and `back` are separate response variables. The term
`(1|p|fosternest)` indicates a varying intercept over `fosternest`. By writing
`|p|` in between we indicate that all varying effects of `fosternest` should be
modeled as correlated. This makes sense since we actually have two model parts,
one for `tarsus` and one for `back`. The indicator `p` is arbitrary and can be
replaced by other symbols that comes into your mind (for details about the
multilevel syntax of **brms**, see `help("brmsformula")` and
`vignette("brms_multilevel")`). Similarly, the term `(1|q|dam)` indicates
correlated varying effects of the genetic mother of the chicks. Alternatively,
we could have also modeled the genetic similarities through pedigrees and
corresponding relatedness matrices, but this is not the focus of this vignette
(please see `vignette("brms_phylogenetics")`). The model results are readily
summarized via

```{r summary1, warning=FALSE}
fit1 <- add_criterion(fit1, "loo")
summary(fit1)
```

The summary output of multivariate models closely resembles those of univariate
models, except that the parameters now have the corresponding response variable
as prefix. Across dams, tarsus length and back color seem to be negatively
correlated, while across fosternests the opposite is true. This indicates
differential effects of genetic and environmental factors on these two
characteristics. Further, the small residual correlation `rescor(tarsus, back)`
on the bottom of the output indicates that there is little unmodeled dependency
between tarsus length and back color. Although not necessary at this point, we
have already computed and stored the LOO information criterion of `fit1`, which
we will use for model comparisons. Next, let's take a look at some
posterior-predictive checks, which give us a first impression of the model fit.

```{r pp_check1, message=FALSE}
pp_check(fit1, resp = "tarsus")
pp_check(fit1, resp = "back")
```

This looks pretty solid, but we notice a slight unmodeled left skewness in the
distribution of `tarsus`. We will come back to this later on. Next, we want to
investigate how much variation in the response variables can be explained by our
model and we use a Bayesian generalization of the $R^2$ coefficient.

```{r R2_1}
bayes_R2(fit1)
```

Clearly, there is much variation in both animal characteristics that we can not
explain, but apparently we can explain more of the variation in tarsus length
than in back color.

## More Complex Multivariate Models

Now, suppose we only want to control for `sex` in `tarsus` but not in `back` and
vice versa for `hatchdate`. Not that this is particular reasonable for the
present example, but it allows us to illustrate how to specify different
formulas for different response variables. We can no longer use `mvbind` syntax
and so we have to use a more verbose approach:

```{r fit2, message=FALSE, warning=FALSE, results='hide'}
bf_tarsus <- bf(tarsus ~ sex + (1 | p | fosternest) + (1 | q | dam))
bf_back <- bf(back ~ hatchdate + (1 | p | fosternest) + (1 | q | dam))
fit2 <- brm(bf_tarsus + bf_back + set_rescor(TRUE),
  data = BTdata, chains = 2, cores = 2
)
```

Note that we have literally *added* the two model parts via the `+` operator,
which is in this case equivalent to writing `mvbf(bf_tarsus, bf_back)`. See
`help("brmsformula")` and `help("mvbrmsformula")` for more details about this
syntax. Again, we summarize the model first.

```{r summary2, warning=FALSE}
fit2 <- add_criterion(fit2, "loo")
summary(fit2)
```

Let's find out, how model fit changed due to excluding certain effects from the
initial model:

```{r loo12}
loo(fit1, fit2)
```

Apparently, there is no noteworthy difference in the model fit. Accordingly, we
do not really need to model `sex` and `hatchdate` for both response variables,
but there is also no harm in including them (so I would probably just include
them).

To give you a glimpse of the capabilities of **brms**' multivariate syntax, we
change our model in various directions at the same time. Remember the slight
left skewness of `tarsus`, which we will now model by using the `skew_normal`
family instead of the `gaussian` family. Since we do not have a multivariate
normal (or student-t) model, anymore, estimating residual correlations is no
longer possible. We make this explicit using the `set_rescor` function. Further,
we investigate if the relationship of `back` and `hatchdate` is really linear as
previously assumed by fitting a non-linear spline of `hatchdate`. On top of it,
we model separate residual variances of `tarsus` for male and female chicks.

```{r fit3, message=FALSE, warning=FALSE, results='hide'}
bf_tarsus <- bf(tarsus ~ sex + (1 | p | fosternest) + (1 | q | dam)) +
  lf(sigma ~ 0 + sex) + skew_normal()
bf_back <- bf(back ~ s(hatchdate) + (1 | p | fosternest) + (1 | q | dam)) +
  gaussian()

fit3 <- brm(
  bf_tarsus + bf_back + set_rescor(FALSE),
  data = BTdata, chains = 2, cores = 2,
  control = list(adapt_delta = 0.95)
)
```

Again, we summarize the model and look at some posterior-predictive checks.

```{r summary3, warning=FALSE}
fit3 <- add_criterion(fit3, "loo")
summary(fit3)
```

We see that the (log) residual standard deviation of `tarsus` is somewhat larger
for chicks whose sex could not be identified as compared to male or female
chicks. Further, we see from the negative `alpha` (skewness) parameter of
`tarsus` that the residuals are indeed slightly left-skewed. Lastly, running

```{r me3}
conditional_effects(fit3, "hatchdate", resp = "back")
```

reveals a non-linear relationship of `hatchdate` on the `back` color, which
seems to change in waves over the course of the hatch dates.

There are many more modeling options for multivariate models, which are not
discussed in this vignette. Examples include autocorrelation structures,
Gaussian processes, or explicit non-linear predictors (e.g., see
`help("brmsformula")` or `vignette("brms_multilevel")`). In fact, nearly all the
flexibility of univariate models is retained in multivariate models.

## References

Hadfield JD, Nutall A, Osorio D, Owens IPF (2007). Testing the phenotypic
gambit: phenotypic, genetic and environmental correlations of colour.
*Journal of Evolutionary Biology*, 20(2), 549-557.
