---
title: "Define Custom Response Distributions with brms"
author: "Paul Bürkner"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: yes
vignette: >
  %\VignetteIndexEntry{Define Custom Response Distributions with brms}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
params:
  EVAL: !r identical(Sys.getenv("NOT_CRAN"), "true")
---

```{r, SETTINGS-knitr, include=FALSE}
stopifnot(require(knitr))
options(width = 90)
knit_hooks$set(pngquant = knitr::hook_pngquant)
opts_chunk$set(
  comment = NA,
  message = FALSE,
  warning = FALSE,
  eval = if (isTRUE(exists("params"))) params$EVAL else FALSE,
  dev = "ragg_png",
  dpi = 72,
  fig.retina = 1.5,
  fig.asp = 0.8,
  fig.width = 5,
  out.width = "60%",
  fig.align = "center",
  pngquant = "--speed=1 --quality=50"
)
library(brms)
ggplot2::theme_set(theme_default())
```

## Introduction

The **brms** package comes with a lot of built-in response distributions --
usually called *families* in R -- to specify among others linear, count data,
survival, response times, or ordinal models (see `help(brmsfamily)` for an
overview). Despite supporting over two dozen families, there is still a long
list of distributions, which are not natively supported. The present vignette
will explain how to specify such *custom families* in **brms**. By doing that,
users can benefit from the modeling flexibility and post-processing options of
**brms** even when using self-defined response distributions.
If you have built a custom family that you want to make available to other
users, you can submit a pull request to this
[GitHub repository](https://github.com/paul-buerkner/custom-brms-families).

## A Case Study

As a case study, we will use the `cbpp` data of the **lme4** package, which
describes the development of the CBPP disease of cattle in Africa. The data set
contains four variables: `period` (the time period), `herd` (a factor
identifying the cattle herd), `incidence` (number of new disease cases for a
given herd and time period), as well as `size` (the herd size at the beginning
of a given time period).

```{r cbpp}
data("cbpp", package = "lme4")
head(cbpp)
```

In a first step, we will be predicting `incidence` using a simple binomial
model, which will serve as our baseline model. For observed number of events $y$
(`incidence` in our case) and total number of trials $T$ (`size`), the
probability mass function of the binomial distribution is defined as

$$
P(y | T, p) = \binom{T}{y} p^{y} (1 - p)^{N-y}
$$

where $p$ is the event probability. In the classical binomial model, we will
directly predict $p$ on the logit-scale, which means that for each observation
$i$ we compute the success probability $p_i$ as

$$
p_i = \frac{\exp(\eta_i)}{1 + \exp(\eta_i)}
$$

where $\eta_i$ is the linear predictor term of observation $i$ (see
`vignette("brms_overview")` for more details on linear predictors in **brms**).
Predicting `incidence` by `period` and a varying intercept of `herd` is straight
forward in **brms**:

```{r fit1, results='hide'}
fit1 <- brm(incidence | trials(size) ~ period + (1|herd),
            data = cbpp, family = binomial())
```

In the summary output, we see that the incidence probability varies
substantially over herds, but reduces over the course of the time as indicated
by the negative coefficients of `period`.

```{r fit1_summary}
summary(fit1)
```

A drawback of the binomial model is that -- after taking into account the linear
predictor -- its variance is fixed to $\text{Var}(y_i) = T_i p_i (1 - p_i)$. All
variance exceeding this value cannot be not taken into account by the model.
There are multiple ways of dealing with this so called *overdispersion* and the
solution described below will serve as an illustrative example of how to define
custom families in **brms**.


## The Beta-Binomial Distribution

The *beta-binomial* model is a generalization of the *binomial* model
with an additional parameter to account for overdispersion. In the beta-binomial
model, we do not predict the binomial probability $p_i$ directly, but assume it
to be beta distributed with hyperparameters $\alpha > 0$ and $\beta > 0$:

$$
p_i \sim \text{Beta}(\alpha_i, \beta_i)
$$

The $\alpha$ and $\beta$ parameters are both hard to interpret and generally
not recommended for use in regression models. Thus, we will apply a different
parameterization with parameters $\mu \in [0, 1]$ and $\phi > 0$, which we will
call $\text{Beta2}$:

$$
\text{Beta2}(\mu, \phi) = \text{Beta}(\mu \phi, (1-\mu) \phi)
$$

The parameters $\mu$ and $\phi$ specify the mean and precision parameter,
respectively. By defining

$$
\mu_i = \frac{\exp(\eta_i)}{1 + \exp(\eta_i)}
$$

we still predict the expected probability by means of our transformed linear
predictor (as in the original binomial model), but account for potential
overdispersion via the parameter $\phi$.


## Fitting Custom Family Models

The beta-binomial distribution is natively supported in **brms** nowadays, but
we will still use it as an example to define it ourselves via the
`custom_family` function. This function requires the family's name, the names of
its parameters (`mu` and `phi` in our case), corresponding link functions (only
applied if parameters are predicted), their theoretical lower and upper bounds
(only applied if parameters are not predicted), information on whether the
distribution is discrete or continuous, and finally, whether additional
non-parameter variables need to be passed to the distribution. For our
beta-binomial example, this results in the following custom family:

```{r beta_binomial2}
beta_binomial2 <- custom_family(
  "beta_binomial2", dpars = c("mu", "phi"),
  links = c("logit", "log"),
  lb = c(0, 0), ub = c(1, NA),
  type = "int", vars = "vint1[n]"
)
```

The name `vint1` for the variable containing the number of trials is not chosen
arbitrarily as we will see below. Next, we have to provide the relevant **Stan**
functions if the distribution is not defined in **Stan** itself. For the
`beta_binomial2` distribution, this is straight forward since the ordinal
`beta_binomial` distribution is already implemented.

```{r stan_funs}
stan_funs <- "
  real beta_binomial2_lpmf(int y, real mu, real phi, int T) {
    return beta_binomial_lpmf(y | T, mu * phi, (1 - mu) * phi);
  }
  int beta_binomial2_rng(real mu, real phi, int T) {
    return beta_binomial_rng(T, mu * phi, (1 - mu) * phi);
  }
"
```

For the model fitting, we will only need `beta_binomial2_lpmf`, but
`beta_binomial2_rng` will come in handy when it comes to post-processing.
We define:

```{r stanvars}
stanvars <- stanvar(scode = stan_funs, block = "functions")
```

To provide information about the number of trials (an integer variable), we are
going to use the addition argument `vint()`, which can only be used in custom
families. Similarly, if we needed to include additional vectors of real data, we
would use `vreal()`. Actually, for this particular example, we could more
elegantly apply the addition argument `trials()` instead of `vint()`as in the
basic binomial model. However, since the present vignette is meant to give a
general overview of the topic, we will go with the more general method.

We now have all components together to fit our custom beta-binomial model:

```{r fit2, results='hide'}
fit2 <- brm(
  incidence | vint(size) ~ period + (1|herd), data = cbpp,
  family = beta_binomial2, stanvars = stanvars
)
```

The summary output reveals that the uncertainty in the coefficients of `period`
is somewhat larger than in the basic binomial model, which is the result of
including the overdispersion parameter `phi` in the model. Apart from that, the
results looks pretty similar.

```{r summary_fit2}
summary(fit2)
```


## Post-Processing Custom Family Models

Some post-processing methods such as `summary` or `plot` work out of the box for
custom family models. However, there are three particularly important methods,
which require additional input by the user. These are `posterior_epred`,
`posterior_predict` and `log_lik` computing predicted mean values, predicted
response values, and log-likelihood values, respectively. They are not only
relevant for their own sake, but also provide the basis of many other
post-processing methods. For instance, we may be interested in comparing the fit
of the binomial model with that of the beta-binomial model by means of
approximate leave-one-out cross-validation implemented in method `loo`, which in
turn requires `log_lik` to be working.

The `log_lik` function of a family should be named `log_lik_<family-name>` and
have the two arguments `i` (indicating observations) and `prep`. You don't have
to worry too much about how `prep` is created (if you are interested, check
out the `prepare_predictions` function). Instead, all you need to know is
that parameters are stored in slot `dpars` and data are stored in slot `data`.
Generally, parameters take on the form of a $S \times N$ matrix (with $S =$
number of posterior draws and $N =$ number of observations) if they are
predicted (as is `mu` in our example) and a vector of size $N$ if the are not
predicted (as is `phi`).

We could define the complete log-likelihood function in R directly, or we can
expose the self-defined **Stan** functions and apply them. The latter approach
is usually more convenient, but the former is more stable and the only option
when implementing custom families in other R packages building upon **brms**.
For the purpose of the present vignette, we will go with the latter approach.

```{r}
expose_functions(fit2, vectorize = TRUE)
```

and define the required `log_lik` functions with a few lines of code.

```{r log_lik}
log_lik_beta_binomial2 <- function(i, prep) {
  mu <- brms::get_dpar(prep, "mu", i = i)
  phi <- brms::get_dpar(prep, "phi", i = i)
  trials <- prep$data$vint1[i]
  y <- prep$data$Y[i]
  beta_binomial2_lpmf(y, mu, phi, trials)
}
```

The `get_dpar` function will do the necessary transformations to handle both
the case when the distributional parameters are predicted separately for each
row and when they are the same for the whole fit.

With that being done, all of the post-processing methods requiring `log_lik`
will work as well. For instance, model comparison can simply be performed via

```{r loo}
loo(fit1, fit2)
```

Since larger `ELPD` values indicate better fit, we see that the beta-binomial
model fits somewhat better, although the corresponding standard error reveals
that the difference is not that substantial.

Next, we will define the function necessary for the `posterior_predict` method.
Historically, custom `posterior_predict` methods only needed to draw random
responses. That style remains supported for the default
`output = "random"` and is enough for posterior-predictive checks:

```{r posterior_predict_old}
posterior_predict_beta_binomial2 <- function(i, prep, ...) {
  mu <- brms::get_dpar(prep, "mu", i = i)
  phi <- brms::get_dpar(prep, "phi", i = i)
  trials <- prep$data$vint1[i]
  beta_binomial2_rng(mu, phi, trials)
}
```

```{r pp_check}
pp_check(fit2)
```

Besides random draws, `posterior_predict` can also return related quantities
such as CDF values (`"probability"`), randomized PIT values (`"pit"`),
densities (`"density"`), or quantiles (`"quantile"`). See the *Details*
section of `help("posterior_predict.brmsfit")` for a full description of
these output types. Custom families only support these modes if their
`posterior_predict` method implements them; otherwise only `"random"` is
available. Methods that do not accept an `output` argument will error when
another mode is requested.

To support the additional modes, accept an `output` argument along with the
usual options such as `q`, `p`, `lower.tail`, `log.p`, and `log` (See the 
*Details* section of `help("posterior_predict.brmsfit")`). For
`beta_binomial2`, the R-side distribution functions of the built-in
`beta_binomial` family apply directly, since both use the same
parameterization:

```{r posterior_predict}
posterior_predict_beta_binomial2 <- function(i, prep, ..., output = "random",
                                             q = NULL, p = NULL,
                                             lower.tail = TRUE, log.p = FALSE,
                                             log = FALSE) {
  mu <- brms::get_dpar(prep, "mu", i = i)
  phi <- brms::get_dpar(prep, "phi", i = i)
  trials <- prep$data$vint1[i]
  if (is.null(q)) {
    q <- prep$data$Y[i]
  }
  switch(
    output,
    "random" = rbeta_binomial(
      n = prep$ndraws, size = trials, mu = mu, phi = phi
    ),
    "probability" = pbeta_binomial(
      q = q, size = trials, mu = mu, phi = phi,
      lower.tail = lower.tail, log.p = log.p
    ),
    "pit" = {
      # randomized PIT for discrete data:
      # F(y-1) + V * [F(y) - F(y-1)] with V ~ Unif(0,1)
      Fy <- pbeta_binomial(q = q, size = trials, mu = mu, phi = phi)
      Fym1 <- pbeta_binomial(q = q - 1, size = trials, mu = mu, phi = phi)
      out <- Fym1 + runif(length(Fy)) * (Fy - Fym1)
      if (!lower.tail) {
        out <- 1 - out
      }
      if (log.p) {
        out <- log(out)
      }
      out
    },
    "density" = dbeta_binomial(
      x = q, size = trials, mu = mu, phi = phi, log = log
    ),
    "quantile" = {
      if (is.null(p)) {
        stop("'p' must be specified when output = 'quantile'.")
      }
      qbeta_binomial(
        p = p, size = trials, mu = mu, phi = phi,
        lower.tail = lower.tail, log.p = log.p
      )
    },
    stop("Unsupported output '", output, "'.")
  )
}
```

The signature places `...` before `output` so that the same function can also
be passed directly to `custom_family(posterior_predict = ...)`, which requires
the first three arguments to be `i`, `prep`, and `...`. With the extended
method in place, we can evaluate the predictive CDF, randomized PIT values,
or extract predictive quantiles:

```{r pp_output_examples}
# predictive CDF at the observed responses
pp_cdf <- posterior_predict(fit2, output = "probability")

# posterior predictive median per observation
pp_median <- posterior_predict(fit2, output = "quantile", p = 0.5)
```

```{r pp_pit_ecdf}
# randomized PIT values (preferred for discrete calibration checks)
pp_pit <- posterior_predict(fit2, output = "pit")

plot(ecdf(colMeans(pp_pit)), 
     xlab = "PIT", ylab = "ECDF", 
     main = NULL, xlim = c(0, 1))
abline(0, 1, lty = 2)
```

When defining the `posterior_epred` function, you have to keep in mind that it
has only a `prep` argument and should compute the mean response values for all
observations at once. Since the mean of the beta-binomial distribution is
$\text{E}(y) = \mu T$ definition of the corresponding `posterior_epred` function
is not too complicated, but we need to get the dimension of parameters and data
in line.

```{r posterior_epred}
posterior_epred_beta_binomial2 <- function(prep) {
  mu <- brms::get_dpar(prep, "mu")
  trials <- prep$data$vint1
  trials <- matrix(trials, nrow = nrow(mu), ncol = ncol(mu), byrow = TRUE)
  mu * trials
}
```

A post-processing method relying directly on `posterior_epred` is
`conditional_effects`, which allows to visualize effects of predictors.

```{r conditional_effects}
conditional_effects(fit2, conditions = data.frame(size = 1))
```

For ease of interpretation we have set `size` to 1 so that the y-axis of the
above plot indicates probabilities.


## Turning a Custom Family into a Native Family

Family functions built natively into **brms** are safer to use and more
convenient, as they require much less user input. If you think that your custom
family is general enough to be useful to other users, please feel free to open
an issue on [GitHub](https://github.com/paul-buerkner/brms/issues) so that we
can discuss all the details. Provided that we agree it makes sense to implement
your family natively in brms, the following steps are required (`foo` is a
placeholder for the family name):

* In `family-lists.R`, add function `.family_foo` which should contain basic
information about your family (you will find lots of examples for other families
there).
* In `families.R`, add family function `foo` which should be a simple wrapper
around `.brmsfamily`.
* In `stan-likelihood.R`, add function `stan_log_lik_foo` which provides the
likelihood of the family in Stan language.
* If necessary, add self-defined Stan functions in separate files under
`inst/chunks`.
* Add functions `posterior_predict_foo`, `posterior_epred_foo` and `log_lik_foo`
to `posterior_predict.R`, `posterior_epred.R` and `log_lik.R`, respectively.
* If necessary, add distribution functions to `distributions.R`.
