---
title: "Cheatsheet: Distribution outputs in posterior_predict()"
author: "brms team"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: yes
vignette: >
  %\VignetteIndexEntry{Cheatsheet: Distribution outputs in posterior_predict()}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
params:
  EVAL: TRUE #!r identical(Sys.getenv("NOT_CRAN"), "true")
---

```{r setup, include=FALSE}
stopifnot(require(knitr))
knit_hooks$set(pngquant = knitr::hook_pngquant)
opts_chunk$set(
  comment = NA,
  message = FALSE,
  warning = FALSE,
  eval = if (isTRUE(exists("params"))) params$EVAL else FALSE,
  dev = "ragg_png",
  dpi = 96,
  fig.retina = 1.5,
  fig.align = "center",
  out.width = "100%",
  pngquant = "--speed=1 --quality=50"
)
library(ggplot2)
devtools::load_all()
if (!requireNamespace("extraDistr", quietly = TRUE)) {
  stop("Package 'extraDistr' is required to evaluate this vignette.")
}
has_betareg <- requireNamespace("betareg", quietly = TRUE)
has_mnormt <- requireNamespace("mnormt", quietly = TRUE)
theme_set(theme_minimal(base_size = 9))
set.seed(4026)
# Defined here so knitr can resolve fig.height even when later chunks
# are skipped (eval = FALSE on CRAN).
n_cols <- 5L
fig_height <- 10.5
```

This cheatsheet summarizes distribution families that currently support **all**
`posterior_predict()` output modes:

- `random`
- `probability`
- `pit`
- `density`
- `quantile`

It is not an exhaustive list of every family that accepts the `output`
argument: some families only support `random`, and a few do not support
posterior predictive draws at all (see [Not fully supported](#not-fully-supported)).

Ordinal families (`cumulative`, `sratio`, `cratio`, `acat`) route through the
shared `posterior_predict_ordinal()` backend.

```{r data-helpers, include=FALSE}
# Inverse-CDF RNG for families without an exported r*() helper
r_from_q <- function(qfun, n, ...) {
  qfun(runif(n), ...)
}

ord_thres <- matrix(c(-1, 0, 1), nrow = 1)
cat_eta <- matrix(c(0, 0.4, -0.2), nrow = 1)

# hurdle_cumulative q/r treat length(p)/rows as draws, not query points
q_hurdle_cumulative_curve <- function(p, hu = 0.25) {
  vapply(
    p,
    function(pi) {
      qhurdle_cumulative(
        pi, eta = 0, thres = ord_thres, hu = hu, link = "logit"
      )
    },
    numeric(1)
  )
}

r_hurdle_cumulative_n <- function(n, hu = 0.25) {
  eta <- rep(0, n)
  thres <- ord_thres[rep(1, n), , drop = FALSE]
  rhurdle_cumulative(
    n, eta = eta, thres = thres, hu = rep(hu, n), link = "logit"
  )
}

r_categorical_n <- function(n) {
  rcategorical(n, eta = cat_eta[rep(1, n), , drop = FALSE])
}

specs <- list(
  # ---- continuous ----
  list(
    family = "gaussian", type = "continuous", x = seq(-4, 4, length.out = 200),
    p = seq(0.01, 0.99, length.out = 200),
    d = function(x) dnorm(x, mean = 0, sd = 1),
    cdf = function(q) pnorm(q, mean = 0, sd = 1),
    qf = function(p) qnorm(p, mean = 0, sd = 1),
    r = function(n) rnorm(n, mean = 0, sd = 1)
  ),
  list(
    family = "student", type = "continuous", x = seq(-6, 6, length.out = 200),
    p = seq(0.01, 0.99, length.out = 200),
    d = function(x) dstudent_t(x, df = 7, mu = 0, sigma = 1),
    cdf = function(q) pstudent_t(q, df = 7, mu = 0, sigma = 1),
    qf = function(p) qstudent_t(p, df = 7, mu = 0, sigma = 1),
    r = function(n) rstudent_t(n, df = 7, mu = 0, sigma = 1)
  ),
  list(
    family = "lognormal", type = "continuous", x = seq(0.05, 8, length.out = 200),
    p = seq(0.01, 0.99, length.out = 200),
    d = function(x) dlnorm(x, meanlog = 0, sdlog = 1),
    cdf = function(q) plnorm(q, meanlog = 0, sdlog = 1),
    qf = function(p) qlnorm(p, meanlog = 0, sdlog = 1),
    r = function(n) rlnorm(n, meanlog = 0, sdlog = 1)
  ),
  list(
    family = "shifted_lognormal", type = "continuous",
    x = seq(0.35, 8, length.out = 200),
    p = seq(0.01, 0.99, length.out = 200),
    d = function(x) dshifted_lnorm(x, meanlog = 0, sdlog = 1, shift = 0.3),
    cdf = function(q) pshifted_lnorm(q, meanlog = 0, sdlog = 1, shift = 0.3),
    qf = function(p) qshifted_lnorm(p, meanlog = 0, sdlog = 1, shift = 0.3),
    r = function(n) rshifted_lnorm(n, meanlog = 0, sdlog = 1, shift = 0.3)
  ),
  list(
    family = "exponential", type = "continuous", x = seq(0.01, 6, length.out = 200),
    p = seq(0.01, 0.99, length.out = 200),
    d = function(x) dexp(x, rate = 1),
    cdf = function(q) pexp(q, rate = 1),
    qf = function(p) qexp(p, rate = 1),
    r = function(n) rexp(n, rate = 1)
  ),
  list(
    family = "gamma", type = "continuous", x = seq(0.01, 10, length.out = 200),
    p = seq(0.01, 0.99, length.out = 200),
    d = function(x) dgamma(x, shape = 2, scale = 1),
    cdf = function(q) pgamma(q, shape = 2, scale = 1),
    qf = function(p) qgamma(p, shape = 2, scale = 1),
    r = function(n) rgamma(n, shape = 2, scale = 1)
  ),
  list(
    family = "weibull", type = "continuous", x = seq(0.01, 4, length.out = 200),
    p = seq(0.01, 0.99, length.out = 200),
    d = function(x) dweibull(x, shape = 2, scale = 1),
    cdf = function(q) pweibull(q, shape = 2, scale = 1),
    qf = function(p) qweibull(p, shape = 2, scale = 1),
    r = function(n) rweibull(n, shape = 2, scale = 1)
  ),
  list(
    family = "frechet", type = "continuous", x = seq(0.2, 6, length.out = 200),
    p = seq(0.01, 0.99, length.out = 200),
    d = function(x) dfrechet(x, scale = 1, shape = 3),
    cdf = function(q) pfrechet(q, scale = 1, shape = 3),
    qf = function(p) qfrechet(p, scale = 1, shape = 3),
    r = function(n) rfrechet(n, scale = 1, shape = 3)
  ),
  list(
    family = "gen_extreme_value", type = "continuous",
    x = seq(-3, 6, length.out = 200),
    p = seq(0.01, 0.99, length.out = 200),
    d = function(x) dgen_extreme_value(x, mu = 0, sigma = 1, xi = 0.1),
    cdf = function(q) pgen_extreme_value(q, mu = 0, sigma = 1, xi = 0.1),
    qf = function(p) qgen_extreme_value(p, mu = 0, sigma = 1, xi = 0.1),
    r = function(n) rgen_extreme_value(n, mu = 0, sigma = 1, xi = 0.1)
  ),
  list(
    family = "inverse.gaussian", type = "continuous",
    x = seq(0.05, 4, length.out = 200),
    p = seq(0.01, 0.99, length.out = 200),
    d = function(x) dinv_gaussian(x, mu = 1, shape = 2),
    cdf = function(q) pinv_gaussian(q, mu = 1, shape = 2),
    qf = function(p) qinv_gaussian(p, mu = 1, shape = 2),
    r = function(n) rinv_gaussian(n, mu = 1, shape = 2)
  ),
  list(
    family = "exgaussian", type = "continuous", x = seq(-2, 8, length.out = 200),
    p = seq(0.01, 0.99, length.out = 200),
    d = function(x) dexgaussian(x, mu = 0, sigma = 1, beta = 1),
    cdf = function(q) pexgaussian(q, mu = 0, sigma = 1, beta = 1),
    qf = function(p) qexgaussian(p, mu = 0, sigma = 1, beta = 1),
    r = function(n) rexgaussian(n, mu = 0, sigma = 1, beta = 1)
  ),
  list(
    family = "asym_laplace", type = "continuous",
    x = seq(-4, 4, length.out = 200),
    p = seq(0.01, 0.99, length.out = 200),
    d = function(x) dasym_laplace(x, mu = 0, sigma = 1, quantile = 0.5),
    cdf = function(q) pasym_laplace(q, mu = 0, sigma = 1, quantile = 0.5),
    qf = function(p) qasym_laplace(p, mu = 0, sigma = 1, quantile = 0.5),
    r = function(n) rasym_laplace(n, mu = 0, sigma = 1, quantile = 0.5)
  ),
  list(
    family = "beta", type = "continuous", x = seq(0.01, 0.99, length.out = 200),
    p = seq(0.01, 0.99, length.out = 200),
    d = function(x) dbeta(x, shape1 = 2, shape2 = 3),
    cdf = function(q) pbeta(q, shape1 = 2, shape2 = 3),
    qf = function(p) qbeta(p, shape1 = 2, shape2 = 3),
    r = function(n) rbeta(n, shape1 = 2, shape2 = 3)
  ),
  list(
    family = "von_mises", type = "continuous",
    x = seq(-pi, pi, length.out = 200),
    p = seq(0.01, 0.99, length.out = 200),
    d = function(x) dvon_mises(x, mu = 0, kappa = 2),
    cdf = function(q) pvon_mises(q, mu = 0, kappa = 2),
    qf = function(p) qvon_mises(p, mu = 0, kappa = 2),
    r = function(n) rvon_mises(n, mu = 0, kappa = 2)
  ),
  # ---- discrete ----
  list(
    family = "bernoulli", type = "discrete", x = 0:1,
    p = seq(0.01, 0.99, length.out = 120),
    d = function(x) dbinom(x, size = 1, prob = 0.4),
    cdf = function(q) pbinom(q, size = 1, prob = 0.4),
    qf = function(p) qbinom(p, size = 1, prob = 0.4),
    r = function(n) rbinom(n, size = 1, prob = 0.4)
  ),
  list(
    family = "binomial", type = "discrete", x = 0:20,
    p = seq(0.01, 0.99, length.out = 120),
    d = function(x) dbinom(x, size = 20, prob = 0.4),
    cdf = function(q) pbinom(q, size = 20, prob = 0.4),
    qf = function(p) qbinom(p, size = 20, prob = 0.4),
    r = function(n) rbinom(n, size = 20, prob = 0.4)
  ),
  list(
    family = "beta_binomial", type = "discrete", x = 0:20,
    p = seq(0.01, 0.99, length.out = 120),
    d = function(x) dbeta_binomial(x, size = 20, mu = 0.4, phi = 8),
    cdf = function(q) pbeta_binomial(q, size = 20, mu = 0.4, phi = 8),
    qf = function(p) qbeta_binomial(p, size = 20, mu = 0.4, phi = 8),
    r = function(n) rbeta_binomial(n, size = 20, mu = 0.4, phi = 8)
  ),
  list(
    family = "poisson", type = "discrete", x = 0:18,
    p = seq(0.01, 0.99, length.out = 120),
    d = function(x) dpois(x, lambda = 5),
    cdf = function(q) ppois(q, lambda = 5),
    qf = function(p) qpois(p, lambda = 5),
    r = function(n) rpois(n, lambda = 5)
  ),
  list(
    family = "negbinomial", type = "discrete", x = 0:24,
    p = seq(0.01, 0.99, length.out = 120),
    d = function(x) dnbinom(x, mu = 5, size = 2),
    cdf = function(q) pnbinom(q, mu = 5, size = 2),
    qf = function(p) qnbinom(p, mu = 5, size = 2),
    r = function(n) rnbinom(n, mu = 5, size = 2)
  ),
  list(
    family = "negbinomial2", type = "discrete", x = 0:24,
    p = seq(0.01, 0.99, length.out = 120),
    d = function(x) dnbinom(x, mu = 5, size = 1 / 0.6),
    cdf = function(q) pnbinom(q, mu = 5, size = 1 / 0.6),
    qf = function(p) qnbinom(p, mu = 5, size = 1 / 0.6),
    r = function(n) rnbinom(n, mu = 5, size = 1 / 0.6)
  ),
  list(
    family = "geometric", type = "discrete", x = 0:20,
    p = seq(0.01, 0.99, length.out = 120),
    d = function(x) dnbinom(x, mu = 3, size = 1),
    cdf = function(q) pnbinom(q, mu = 3, size = 1),
    qf = function(p) qnbinom(p, mu = 3, size = 1),
    r = function(n) rnbinom(n, mu = 3, size = 1)
  ),
  list(
    family = "discrete_weibull", type = "discrete", x = 0:15,
    p = seq(0.01, 0.99, length.out = 120),
    d = function(x) ddiscrete_weibull(x, mu = 0.7, shape = 1.5),
    cdf = function(q) pdiscrete_weibull(q, mu = 0.7, shape = 1.5),
    qf = function(p) qdiscrete_weibull(p, mu = 0.7, shape = 1.5),
    r = function(n) rdiscrete_weibull(n, mu = 0.7, shape = 1.5)
  ),
  list(
    family = "com_poisson", type = "discrete", x = 0:12,
    p = seq(0.01, 0.99, length.out = 80),
    d = function(x) dcom_poisson(x, mu = 2, shape = 0.8),
    cdf = function(q) pcom_poisson(q, mu = 2, shape = 0.8),
    qf = function(p) qcom_poisson(p, mu = 2, shape = 0.8),
    r = function(n) rcom_poisson(n, mu = 2, shape = 0.8)
  ),
  # ---- hurdle ----
  list(
    family = "hurdle_poisson", type = "discrete", x = 0:18,
    p = seq(0.01, 0.99, length.out = 120),
    d = function(x) dhurdle_poisson(x, lambda = 5, hu = 0.25),
    cdf = function(q) phurdle_poisson(q, lambda = 5, hu = 0.25),
    qf = function(p) qhurdle_poisson(p, lambda = 5, hu = 0.25),
    r = function(n) rhurdle_poisson(n, lambda = 5, hu = 0.25)
  ),
  list(
    family = "hurdle_negbinomial", type = "discrete", x = 0:24,
    p = seq(0.01, 0.99, length.out = 120),
    d = function(x) dhurdle_negbinomial(x, mu = 5, shape = 2, hu = 0.25),
    cdf = function(q) phurdle_negbinomial(q, mu = 5, shape = 2, hu = 0.25),
    qf = function(p) qhurdle_negbinomial(p, mu = 5, shape = 2, hu = 0.25),
    r = function(n) rhurdle_negbinomial(n, mu = 5, shape = 2, hu = 0.25)
  ),
  list(
    family = "hurdle_gamma", type = "mixture", x = seq(0, 10, length.out = 200),
    p = seq(0.01, 0.99, length.out = 200),
    d = function(x) dhurdle_gamma(x, shape = 2, scale = 1, hu = 0.25),
    cdf = function(q) phurdle_gamma(q, shape = 2, scale = 1, hu = 0.25),
    qf = function(p) qhurdle_gamma(p, shape = 2, scale = 1, hu = 0.25),
    r = function(n) rhurdle_gamma(n, shape = 2, scale = 1, hu = 0.25)
  ),
  list(
    family = "hurdle_lognormal", type = "mixture",
    x = seq(0, 8, length.out = 200),
    p = seq(0.01, 0.99, length.out = 200),
    d = function(x) dhurdle_lognormal(x, mu = 0, sigma = 1, hu = 0.25),
    cdf = function(q) phurdle_lognormal(q, mu = 0, sigma = 1, hu = 0.25),
    qf = function(p) qhurdle_lognormal(p, mu = 0, sigma = 1, hu = 0.25),
    r = function(n) rhurdle_lognormal(n, mu = 0, sigma = 1, hu = 0.25)
  ),
  list(
    family = "hurdle_cumulative", type = "discrete", x = 0:4,
    p = seq(0.01, 0.99, length.out = 120),
    d = function(x) {
      as.numeric(dhurdle_cumulative(
        x, eta = 0, thres = ord_thres, hu = 0.25, link = "logit"
      ))
    },
    cdf = function(q) {
      as.numeric(phurdle_cumulative(
        q, eta = 0, thres = ord_thres, hu = 0.25, link = "logit"
      ))
    },
    qf = function(p) q_hurdle_cumulative_curve(p, hu = 0.25),
    r = function(n) r_hurdle_cumulative_n(n, hu = 0.25)
  ),
  # ---- zero-inflated ----
  list(
    family = "zero_inflated_poisson", type = "discrete", x = 0:18,
    p = seq(0.01, 0.99, length.out = 120),
    d = function(x) dzero_inflated_poisson(x, lambda = 5, zi = 0.25),
    cdf = function(q) pzero_inflated_poisson(q, lambda = 5, zi = 0.25),
    qf = function(p) qzero_inflated_poisson(p, lambda = 5, zi = 0.25),
    r = function(n) r_from_q(qzero_inflated_poisson, n, lambda = 5, zi = 0.25)
  ),
  list(
    family = "zero_inflated_negbinomial", type = "discrete", x = 0:24,
    p = seq(0.01, 0.99, length.out = 120),
    d = function(x) dzero_inflated_negbinomial(x, mu = 5, shape = 2, zi = 0.25),
    cdf = function(q) {
      pzero_inflated_negbinomial(q, mu = 5, shape = 2, zi = 0.25)
    },
    qf = function(p) {
      qzero_inflated_negbinomial(p, mu = 5, shape = 2, zi = 0.25)
    },
    r = function(n) {
      r_from_q(qzero_inflated_negbinomial, n, mu = 5, shape = 2, zi = 0.25)
    }
  ),
  list(
    family = "zero_inflated_binomial", type = "discrete", x = 0:20,
    p = seq(0.01, 0.99, length.out = 120),
    d = function(x) dzero_inflated_binomial(x, size = 20, prob = 0.4, zi = 0.2),
    cdf = function(q) {
      pzero_inflated_binomial(q, size = 20, prob = 0.4, zi = 0.2)
    },
    qf = function(p) {
      qzero_inflated_binomial(p, size = 20, prob = 0.4, zi = 0.2)
    },
    r = function(n) {
      r_from_q(qzero_inflated_binomial, n, size = 20, prob = 0.4, zi = 0.2)
    }
  ),
  list(
    family = "zero_inflated_beta_binomial", type = "discrete", x = 0:20,
    p = seq(0.01, 0.99, length.out = 120),
    d = function(x) {
      dzero_inflated_beta_binomial(x, size = 20, mu = 0.4, phi = 8, zi = 0.2)
    },
    cdf = function(q) {
      pzero_inflated_beta_binomial(q, size = 20, mu = 0.4, phi = 8, zi = 0.2)
    },
    qf = function(p) {
      qzero_inflated_beta_binomial(p, size = 20, mu = 0.4, phi = 8, zi = 0.2)
    },
    r = function(n) {
      r_from_q(
        qzero_inflated_beta_binomial, n,
        size = 20, mu = 0.4, phi = 8, zi = 0.2
      )
    }
  ),
  list(
    family = "zero_inflated_beta", type = "mixture",
    x = seq(0, 1, length.out = 200),
    p = seq(0.01, 0.99, length.out = 200),
    d = function(x) dzero_inflated_beta(x, shape1 = 2, shape2 = 3, zi = 0.25),
    cdf = function(q) pzero_inflated_beta(q, shape1 = 2, shape2 = 3, zi = 0.25),
    qf = function(p) qzero_inflated_beta(p, shape1 = 2, shape2 = 3, zi = 0.25),
    r = function(n) {
      r_from_q(qzero_inflated_beta, n, shape1 = 2, shape2 = 3, zi = 0.25)
    }
  ),
  list(
    family = "zero_inflated_asym_laplace", type = "mixture",
    x = seq(-4, 4, length.out = 200),
    p = seq(0.01, 0.99, length.out = 200),
    d = function(x) {
      dzero_inflated_asym_laplace(
        x, mu = 0, sigma = 1, quantile = 0.5, zi = 0.3
      )
    },
    cdf = function(q) {
      pzero_inflated_asym_laplace(
        q, mu = 0, sigma = 1, quantile = 0.5, zi = 0.3
      )
    },
    qf = function(p) {
      qzero_inflated_asym_laplace(
        p, mu = 0, sigma = 1, quantile = 0.5, zi = 0.3
      )
    },
    r = function(n) {
      r_from_q(
        qzero_inflated_asym_laplace, n,
        mu = 0, sigma = 1, quantile = 0.5, zi = 0.3
      )
    }
  ),
  list(
    family = "zero_one_inflated_beta", type = "mixture",
    x = seq(0, 1, length.out = 200),
    p = seq(0.01, 0.99, length.out = 200),
    d = function(x) {
      dzero_one_inflated_beta(
        x, shape1 = 2, shape2 = 3, zoi = 0.25, coi = 0.4
      )
    },
    cdf = function(q) {
      pzero_one_inflated_beta(
        q, shape1 = 2, shape2 = 3, zoi = 0.25, coi = 0.4
      )
    },
    qf = function(p) {
      qzero_one_inflated_beta(
        p, shape1 = 2, shape2 = 3, zoi = 0.25, coi = 0.4
      )
    },
    r = function(n) {
      r_from_q(
        qzero_one_inflated_beta, n,
        shape1 = 2, shape2 = 3, zoi = 0.25, coi = 0.4
      )
    }
  )
)

# ---- ordinal / categorical ----
ordinal_specs <- lapply(c("cumulative", "sratio", "cratio", "acat"), function(fam) {
  force(fam)
  list(
    family = fam, type = "discrete", x = 1:4,
    p = seq(0.01, 0.99, length.out = 120),
    d = function(x) {
      as.numeric(dordinal(
        x, eta = 0, thres = ord_thres, family = fam, link = "logit"
      ))
    },
    cdf = function(q) {
      as.numeric(pordinal(
        q, eta = 0, thres = ord_thres, family = fam, link = "logit"
      ))
    },
    qf = function(p) {
      as.numeric(qordinal(
        p, eta = 0, thres = ord_thres, family = fam, link = "logit"
      ))
    },
    r = function(n) {
      as.numeric(rordinal(
        n, eta = 0, thres = ord_thres, family = fam, link = "logit"
      ))
    }
  )
})

categorical_spec <- list(
  family = "categorical", type = "discrete", x = 1:3,
  p = seq(0.01, 0.99, length.out = 120),
  d = function(x) as.numeric(dcategorical(x, eta = cat_eta)),
  cdf = function(q) as.numeric(pcategorical(q, eta = cat_eta)),
  qf = function(p) as.numeric(qcategorical(p, eta = cat_eta)),
  r = function(n) as.numeric(r_categorical_n(n))
)

specs <- c(specs, ordinal_specs, list(categorical_spec))

if (has_mnormt) {
  skew_spec <- list(
    family = "skew_normal", type = "continuous",
    x = seq(-3, 5, length.out = 200),
    p = seq(0.01, 0.99, length.out = 200),
    d = function(x) dskew_normal(x, mu = 0, sigma = 1, alpha = 2),
    cdf = function(q) pskew_normal(q, mu = 0, sigma = 1, alpha = 2),
    qf = function(p) qskew_normal(p, mu = 0, sigma = 1, alpha = 2),
    r = function(n) rskew_normal(n, mu = 0, sigma = 1, alpha = 2)
  )
  shifted_idx <- match("shifted_lognormal", vapply(specs, `[[`, "", "family"))
  specs <- append(specs, list(skew_spec), after = shifted_idx)
}

if (has_betareg) {
  xbeta_spec <- list(
    family = "xbeta", type = "continuous",
    x = seq(0.01, 0.99, length.out = 200),
    p = seq(0.01, 0.99, length.out = 200),
    d = function(x) dxbeta(x, mu = 0.5, phi = 5, nu = 1),
    cdf = function(q) pxbeta(q, mu = 0.5, phi = 5, nu = 1),
    qf = function(p) qxbeta(p, mu = 0.5, phi = 5, nu = 1),
    r = function(n) rxbeta(n, mu = 0.5, phi = 5, nu = 1)
  )
  beta_idx <- match("beta", vapply(specs, `[[`, "", "family"))
  specs <- append(specs, list(xbeta_spec), after = beta_idx)
}


# Family -> compute_*/predict_* distribution backend name
family_backends <- c(
  gaussian = "norm",
  student = "student_t",
  lognormal = "lnorm",
  shifted_lognormal = "shifted_lnorm",
  skew_normal = "skew_normal",
  exponential = "exp",
  gamma = "gamma",
  weibull = "weibull",
  frechet = "frechet",
  gen_extreme_value = "gen_extreme_value",
  "inverse.gaussian" = "inv_gaussian",
  exgaussian = "exgaussian",
  asym_laplace = "asym_laplace",
  beta = "beta",
  xbeta = "xbeta",
  von_mises = "von_mises",
  bernoulli = "binom",
  binomial = "binom",
  beta_binomial = "beta_binomial",
  poisson = "pois",
  negbinomial = "nbinom",
  negbinomial2 = "nbinom",
  geometric = "nbinom",
  discrete_weibull = "discrete_weibull",
  com_poisson = "com_poisson",
  hurdle_poisson = "hurdle_poisson",
  hurdle_negbinomial = "hurdle_negbinomial",
  hurdle_gamma = "hurdle_gamma",
  hurdle_lognormal = "hurdle_lognormal",
  hurdle_cumulative = "hurdle_cumulative",
  zero_inflated_poisson = "zero_inflated_poisson",
  zero_inflated_negbinomial = "zero_inflated_negbinomial",
  zero_inflated_binomial = "zero_inflated_binomial",
  zero_inflated_beta_binomial = "zero_inflated_beta_binomial",
  zero_inflated_beta = "zero_inflated_beta",
  zero_inflated_asym_laplace = "zero_inflated_asym_laplace",
  zero_one_inflated_beta = "zero_one_inflated_beta",
  cumulative = "ordinal",
  sratio = "ordinal",
  cratio = "ordinal",
  acat = "ordinal",
  categorical = "categorical"
)

fams <- vapply(specs, `[[`, character(1), "family")
missing_backend <- setdiff(fams, names(family_backends))
if (length(missing_backend)) {
  stop("Missing family_backends for: ", paste(missing_backend, collapse = ", "))
}
tick <- "✓"
support <- data.frame(
  family = fams,
  distribution_backend = unname(family_backends[fams]),
  random = tick,
  probability = tick,
  pit = tick,
  density = tick,
  quantile = tick,
  stringsAsFactors = FALSE
)

make_output_data <- function(spec) {
  density_df <- data.frame(
    family = spec$family,
    type = spec$type,
    x = spec$x,
    y = as.numeric(spec$d(spec$x))
  )
  # continuous mixture densities may be Inf at atoms; drop for plotting
  density_df$y[!is.finite(density_df$y)] <- NA_real_

  probability_df <- data.frame(
    family = spec$family,
    type = spec$type,
    x = spec$x,
    y = as.numeric(spec$cdf(spec$x))
  )
  quantile_df <- data.frame(
    family = spec$family,
    type = spec$type,
    x = spec$p,
    y = as.numeric(spec$qf(spec$p))
  )

  draws <- as.numeric(spec$r(500))
  ecdf_fun <- ecdf(draws)
  random_df <- data.frame(
    family = spec$family,
    type = spec$type,
    x = spec$x,
    y = ecdf_fun(spec$x)
  )

  pit_draws <- as.numeric(spec$r(400))
  if (identical(spec$type, "continuous")) {
    pit <- as.numeric(spec$cdf(pit_draws))
  } else if (identical(spec$type, "mixture")) {
    # continuous mixtures with atoms: use CDF (atoms appear as spikes)
    pit <- as.numeric(spec$cdf(pit_draws))
  } else {
    u <- runif(length(pit_draws))
    pit <- as.numeric(spec$cdf(pit_draws - 1)) +
      u * (as.numeric(spec$cdf(pit_draws)) - as.numeric(spec$cdf(pit_draws - 1)))
  }
  pit <- pmax(0, pmin(1, pit))
  pit_hist <- hist(pit, breaks = seq(0, 1, by = 0.1), plot = FALSE)
  pit_df <- data.frame(
    family = spec$family,
    midpoint = pit_hist$mids,
    density = pit_hist$density
  )

  list(
    density = density_df,
    probability = probability_df,
    quantile = quantile_df,
    random = random_df,
    pit = pit_df
  )
}

all_data <- lapply(specs, make_output_data)
density_data <- do.call(rbind, lapply(all_data, `[[`, "density"))
probability_data <- do.call(rbind, lapply(all_data, `[[`, "probability"))
quantile_data <- do.call(rbind, lapply(all_data, `[[`, "quantile"))
random_data <- do.call(rbind, lapply(all_data, `[[`, "random"))
pit_data <- do.call(rbind, lapply(all_data, `[[`, "pit"))

n_families <- length(specs)
fig_height <- max(4.2, 1.15 * ceiling(n_families / n_cols))

panel_theme <- theme(
  strip.text = element_text(size = 7),
  axis.title = element_blank(),
  panel.grid.minor = element_blank()
)

# Shared facet + labels wrapper; `layers` is a list of geoms.
thumb_plot <- function(layers, title, ylab, scales = "free_x", xlab = NULL) {
  ggplot() +
    layers +
    facet_wrap(~ family, scales = scales, ncol = n_cols) +
    labs(title = title, x = xlab, y = ylab) +
    panel_theme
}
```


## Supported families (full five-mode support)

The table is generated from the same family specs used for the thumbnails
below (optional families appear only when their Suggested packages are
installed: `skew_normal` needs `mnormt`, `xbeta` needs `betareg`).

```{r support-table}
knitr::kable(support, align = "l")
```

## Not fully supported

These families accept `posterior_predict()` but do **not** implement all five
output modes (or cannot draw from the predictive at all):

| Category | Families | Supported `output` |
|----------|----------|--------------------|
| Multivariate / time / spatial / residual correlation | `gaussian_mv`, `student_mv`, `gaussian_time`, `student_time`, `gaussian_lagsar`, `student_lagsar`, `gaussian_errorsar`, `student_errorsar`, `gaussian_fcor`, `student_fcor` | `random` only |
| Diffusion | `wiener` | `random` only |
| Compositional / multinomial | `multinomial`, `dirichlet_multinomial`, `dirichlet`, `dirichlet2`, `logistic_normal` | `random` only |
| Mixtures of families | `mixture` | `random` only |
| Survival (no predictive draws) | `cox` | none (`posterior_predict()` errors) |
| User-defined | `custom` | `random` by default; other modes only if the custom family implements them |

Truncation is handled via the shared continuous (inverse-CDF) and discrete
(rejection sampling) helpers for all fully supported families, including the
hurdle families.

## Mathematical details

Let \(Y\) be the posterior predictive variable for one posterior draw \(s\), with
distribution \(F_s\) and density/mass \(f_s\). For each observation index \(i\),
`posterior_predict()` returns one value per draw \(s = 1, \dots, S\).

### 1) `output = "random"`

Draw
\[
\tilde y_s \sim F_s.
\]
If truncation bounds are active (\(L, U\)), sampling is from the truncated
distribution:
\[
F_{s,\text{tr}}(y) = \frac{F_s(y) - F_s(L)}{F_s(U) - F_s(L)}, \quad L \le y \le U.
\]

### 2) `output = "probability"`

For a queried value \(q\), return the (possibly truncated) CDF:
\[
p_s(q) = F_s(q)
\]
or, with truncation,
\[
p_s(q) = \frac{F_s(q) - F_s(L)}{F_s(U) - F_s(L)}.
\]
With `lower.tail = FALSE`, return \(1 - p_s(q)\). With `log.p = TRUE`, return
\(\log p_s(q)\).

### 3) `output = "pit"`

- Continuous case: PIT equals the CDF value,
\[
\text{PIT}_s = F_s(y_i).
\]
- Discrete case: randomized PIT (to avoid spikes),
\[
\text{PIT}_s = F_s(y_i - 1) + V_s \left[F_s(y_i) - F_s(y_i - 1)\right], \quad
V_s \sim \text{Uniform}(0,1).
\]

### 4) `output = "density"`

For a queried value \(q\), return
\[
d_s(q) = f_s(q)
\]
or with truncation,
\[
d_s(q) = \frac{f_s(q)}{F_s(U) - F_s(L)} \mathbf{1}\{L \le q \le U\}.
\]
With `log = TRUE`, return \(\log d_s(q)\).

For discrete distributions, this corresponds to PMF values and satisfies
\[
d_s(k) = F_s(k) - F_s(k-1).
\]

### 5) `output = "quantile"`

For queried probability \(p\), return
\[
q_s(p) = F_s^{-1}(p).
\]
With truncation, the internally transformed probability is
\[
p_s^\star = p \left[F_s(U) - F_s(L)\right] + F_s(L),
\]
and then
\[
q_s(p) = F_s^{-1}(p_s^\star).
\]
`lower.tail` and `log.p` are first converted to standard probabilities before
applying these formulas.

### Zero-inflated and hurdle families

For zero-inflated families (parameter \(\pi\), baseline distribution \(G\)):
\[
P(Y=0) = \pi + (1-\pi)G(0), \qquad
P(Y=y>0) = (1-\pi)G(y).
\]

For hurdle families (parameter \(h\), baseline distribution \(G\)):
\[
P(Y=0) = h, \qquad
P(Y=y>0) = (1-h)\frac{G(y)}{1-G(0)}.
\]

These definitions induce the corresponding CDF, density/mass, and quantile
computations used by `probability`, `density`, and `quantile`.

## Random output

Empirical CDF of random draws (`output = "random"`).

```{r random-plot, fig.width=9, fig.height=fig_height}
thumb_plot(
  list(
    geom_point(
      data = subset(random_data, type != "discrete"),
      aes(x = x, y = y),
      size = 0.4, color = "#2c7fb8"
    ),
    geom_point(
      data = subset(random_data, type == "discrete"),
      aes(x = x, y = y),
      size = 1.1, color = "#2c7fb8"
    )
  ),
  title = "Random output (empirical CDF thumbnails)",
  ylab = "ECDF"
)
```

## Probability output

CDF values (`output = "probability"`).

```{r probability-plot, fig.width=9, fig.height=fig_height}
thumb_plot(
  list(
    geom_line(
      data = subset(probability_data, type %in% c("continuous", "mixture")),
      aes(x = x, y = y),
      linewidth = 0.4, color = "#1b9e77"
    ),
    geom_step(
      data = subset(probability_data, type == "discrete"),
      aes(x = x, y = y),
      linewidth = 0.4, color = "#1b9e77"
    )
  ),
  title = "Probability output (CDF thumbnails)",
  ylab = "P(X <= x)"
)
```

## PIT output

PIT histogram (`output = "pit"`). For calibrated models, these are near-uniform.

```{r pit-plot, fig.width=9, fig.height=fig_height}
thumb_plot(
  list(
    geom_col(
      data = pit_data, aes(x = midpoint, y = density),
      width = 0.085, fill = "#7570b3", alpha = 0.75
    ),
    geom_hline(yintercept = 1, linewidth = 0.25, linetype = "dashed"),
    scale_x_continuous(limits = c(0, 1), breaks = c(0, 0.5, 1))
  ),
  title = "PIT output (thumbnail histograms)",
  ylab = "Density",
  scales = "fixed"
)
```

## Density output

PDF/PMF values (`output = "density"`). Point masses of mixture families are
omitted from the continuous density curves.

```{r density-plot, fig.width=9, fig.height=fig_height}
thumb_plot(
  list(
    geom_line(
      data = subset(density_data, type %in% c("continuous", "mixture")),
      aes(x = x, y = y),
      linewidth = 0.45, color = "#d95f02", na.rm = TRUE
    ),
    geom_col(
      data = subset(density_data, type == "discrete"),
      aes(x = x, y = y),
      width = 0.85, fill = "#d95f02", alpha = 0.7
    )
  ),
  title = "Density output (PDF/PMF thumbnails)",
  ylab = "Density / Mass",
  scales = "free"
)
```

## Quantile output

Inverse CDF values (`output = "quantile"`).

```{r quantile-plot, fig.width=9, fig.height=fig_height}
thumb_plot(
  list(
    geom_line(
      data = subset(quantile_data, type %in% c("continuous", "mixture")),
      aes(x = x, y = y),
      linewidth = 0.4, color = "#e7298a"
    ),
    geom_step(
      data = subset(quantile_data, type == "discrete"),
      aes(x = x, y = y),
      linewidth = 0.4, color = "#e7298a"
    )
  ),
  title = "Quantile output (inverse-CDF thumbnails)",
  ylab = "Q(p)",
  xlab = "p",
  scales = "free_y"
)
```

## Notes

- Continuous families have `probability == pit`.
- Discrete and ordinal families use randomized PIT to avoid spikes.
- Mixture families with point masses (hurdle continuous, zero-/zero-one-inflated
  beta and asymmetric Laplace) have atoms that are handled in the CDF/quantile
  helpers; density thumbnails show only the continuous component.
- The parameter values here are illustrative; the output shape changes with
  posterior draws in real model fits.
- `skew_normal` thumbnails require Suggested package `mnormt`; `xbeta`
  thumbnails require Suggested package `betareg`.
