---
title: "Cheatsheet: Distribution outputs in posterior_predict()"
author: "brms team"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: yes
vignette: >
  %\VignetteIndexEntry{Cheatsheet: Distribution outputs in posterior_predict()}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
params:
  EVAL: !r identical(Sys.getenv("NOT_CRAN"), "true")
---

```{r setup, include=FALSE}
stopifnot(require(knitr))
knit_hooks$set(pngquant = knitr::hook_pngquant)
opts_chunk$set(
  comment = NA,
  message = FALSE,
  warning = FALSE,
  eval = if (isTRUE(exists("params"))) params$EVAL else FALSE,
  dev = "ragg_png",
  dpi = 96,
  fig.retina = 1.5,
  fig.align = "center",
  out.width = "100%",
  pngquant = "--speed=1 --quality=50"
)
library(ggplot2)
devtools::load_all()
if (!requireNamespace("extraDistr", quietly = TRUE)) {
  stop("Package 'extraDistr' is required to evaluate this vignette.")
}
theme_set(theme_minimal(base_size = 9))
set.seed(4026)
```

This cheatsheet summarizes the distribution families that currently support all
`posterior_predict()` output modes:

- `random`
- `probability`
- `pit`
- `density`
- `quantile`

## Supported families

```{r support-table}
support <- data.frame(
  family = c(
    "gaussian", "student", "binomial", "beta_binomial",
    "poisson", "negbinomial", "negbinomial2", "zero_inflated_negbinomial"
  ),
  distribution_backend = c(
    "norm", "student_t", "binom", "beta_binomial",
    "pois", "nbinom", "nbinom", "zero_inflated_negbinomial"
  ),
  random = "\u2713",
  probability = "\u2713",
  pit = "\u2713",
  density = "\u2713",
  quantile = "\u2713",
  stringsAsFactors = FALSE
)
knitr::kable(support, align = "l")
```

## Mathematical details

Let \(Y\) be the posterior predictive variable for one posterior draw \(s\), with
distribution \(F_s\) and density/mass \(f_s\). For each observation index \(i\),
`posterior_predict()` returns one value per draw \(s = 1, \dots, S\).

### 1) `output = "random"`

Draw
\[
\tilde y_s \sim F_s.
\]
If truncation bounds are active (\(L, U\)), sampling is from the truncated
distribution:
\[
F_{s,\text{tr}}(y) = \frac{F_s(y) - F_s(L)}{F_s(U) - F_s(L)}, \quad L \le y \le U.
\]

### 2) `output = "probability"`

For a queried value \(q\), return the (possibly truncated) CDF:
\[
p_s(q) = F_s(q)
\]
or, with truncation,
\[
p_s(q) = \frac{F_s(q) - F_s(L)}{F_s(U) - F_s(L)}.
\]
With `lower.tail = FALSE`, return \(1 - p_s(q)\). With `log.p = TRUE`, return
\(\log p_s(q)\).

### 3) `output = "pit"`

- Continuous case: PIT equals the CDF value,
\[
\text{PIT}_s = F_s(y_i).
\]
- Discrete case: randomized PIT (to avoid spikes),
\[
\text{PIT}_s = F_s(y_i - 1) + V_s \left[F_s(y_i) - F_s(y_i - 1)\right], \quad
V_s \sim \text{Uniform}(0,1).
\]

### 4) `output = "density"`

For a queried value \(q\), return
\[
d_s(q) = f_s(q)
\]
or with truncation,
\[
d_s(q) = \frac{f_s(q)}{F_s(U) - F_s(L)} \mathbf{1}\{L \le q \le U\}.
\]
With `log = TRUE`, return \(\log d_s(q)\).

For discrete distributions, this corresponds to PMF values and satisfies
\[
d_s(k) = F_s(k) - F_s(k-1).
\]

### 5) `output = "quantile"`

For queried probability \(p\), return
\[
q_s(p) = F_s^{-1}(p).
\]
With truncation, the internally transformed probability is
\[
p_s^\star = p \left[F_s(U) - F_s(L)\right] + F_s(L),
\]
and then
\[
q_s(p) = F_s^{-1}(p_s^\star).
\]
`lower.tail` and `log.p` are first converted to standard probabilities before
applying these formulas.

### Zero-inflated and hurdle families

For zero-inflated families (parameter \(\pi\), baseline distribution \(G\)):
\[
P(Y=0) = \pi + (1-\pi)G(0), \qquad
P(Y=y>0) = (1-\pi)G(y).
\]

For hurdle families (parameter \(h\), baseline distribution \(G\)):
\[
P(Y=0) = h, \qquad
P(Y=y>0) = (1-h)\frac{G(y)}{1-G(0)}.
\]

These definitions induce the corresponding CDF, density/mass, and quantile
computations used by `probability`, `density`, and `quantile`.

```{r data-helpers, include=FALSE}
specs <- list(
  list(
    family = "gaussian", type = "continuous", x = seq(-4, 4, length.out = 200),
    p = seq(0.01, 0.99, length.out = 200),
    d = function(x) dnorm(x, mean = 0, sd = 1),
    cdf = function(q) pnorm(q, mean = 0, sd = 1),
    qf = function(p) qnorm(p, mean = 0, sd = 1),
    r = function(n) rnorm(n, mean = 0, sd = 1)
  ),
  list(
    family = "student", type = "continuous", x = seq(-6, 6, length.out = 200),
    p = seq(0.01, 0.99, length.out = 200),
    d = function(x) dstudent_t(x, df = 7, mu = 0, sigma = 1),
    cdf = function(q) pstudent_t(q, df = 7, mu = 0, sigma = 1),
    qf = function(p) qstudent_t(p, df = 7, mu = 0, sigma = 1),
    r = function(n) rstudent_t(n, df = 7, mu = 0, sigma = 1)
  ),
  list(
    family = "binomial", type = "discrete", x = 0:20,
    p = seq(0.01, 0.99, length.out = 120),
    d = function(x) dbinom(x, size = 20, prob = 0.4),
    cdf = function(q) pbinom(q, size = 20, prob = 0.4),
    qf = function(p) qbinom(p, size = 20, prob = 0.4),
    r = function(n) rbinom(n, size = 20, prob = 0.4)
  ),
  list(
    family = "beta_binomial", type = "discrete", x = 0:20,
    p = seq(0.01, 0.99, length.out = 120),
    d = function(x) dbeta_binomial(x, size = 20, mu = 0.4, phi = 8),
    cdf = function(q) pbeta_binomial(q, size = 20, mu = 0.4, phi = 8),
    qf = function(p) qbeta_binomial(p, size = 20, mu = 0.4, phi = 8),
    r = function(n) rbeta_binomial(n, size = 20, mu = 0.4, phi = 8)
  ),
  list(
    family = "poisson", type = "discrete", x = 0:18,
    p = seq(0.01, 0.99, length.out = 120),
    d = function(x) dpois(x, lambda = 5),
    cdf = function(q) ppois(q, lambda = 5),
    qf = function(p) qpois(p, lambda = 5),
    r = function(n) rpois(n, lambda = 5)
  ),
  list(
    family = "negbinomial", type = "discrete", x = 0:24,
    p = seq(0.01, 0.99, length.out = 120),
    d = function(x) dnbinom(x, mu = 5, size = 2),
    cdf = function(q) pnbinom(q, mu = 5, size = 2),
    qf = function(p) qnbinom(p, mu = 5, size = 2),
    r = function(n) rnbinom(n, mu = 5, size = 2)
  ),
  list(
    family = "negbinomial2", type = "discrete", x = 0:24,
    p = seq(0.01, 0.99, length.out = 120),
    d = function(x) dnbinom(x, mu = 5, size = 1 / 0.6),
    cdf = function(q) pnbinom(q, mu = 5, size = 1 / 0.6),
    qf = function(p) qnbinom(p, mu = 5, size = 1 / 0.6),
    r = function(n) rnbinom(n, mu = 5, size = 1 / 0.6)
  ),
  list(
    family = "zero_inflated_negbinomial", type = "discrete", x = 0:24,
    p = seq(0.01, 0.99, length.out = 120),
    d = function(x) dzero_inflated_negbinomial(x, mu = 5, shape = 2, zi = 0.25),
    cdf = function(q) pzero_inflated_negbinomial(q, mu = 5, shape = 2, zi = 0.25),
    qf = function(p) qzero_inflated_negbinomial(p, mu = 5, shape = 2, zi = 0.25),
    r = function(n) {
      base <- rnbinom(n, mu = 5, size = 2)
      is_zi <- runif(n) < 0.25
      base[is_zi] <- 0
      base
    }
  )
)

make_output_data <- function(spec) {
  density_df <- data.frame(
    family = spec$family,
    type = spec$type,
    x = spec$x,
    y = spec$d(spec$x)
  )
  probability_df <- data.frame(
    family = spec$family,
    type = spec$type,
    x = spec$x,
    y = spec$cdf(spec$x)
  )
  quantile_df <- data.frame(
    family = spec$family,
    type = spec$type,
    x = spec$p,
    y = spec$qf(spec$p)
  )

  draws <- spec$r(500)
  ecdf_fun <- ecdf(draws)
  random_df <- data.frame(
    family = spec$family,
    type = spec$type,
    x = spec$x,
    y = ecdf_fun(spec$x)
  )

  pit_draws <- spec$r(400)
  if (identical(spec$type, "continuous")) {
    pit <- spec$cdf(pit_draws)
  } else {
    u <- runif(length(pit_draws))
    pit <- spec$cdf(pit_draws - 1) + u * (spec$cdf(pit_draws) - spec$cdf(pit_draws - 1))
  }
  pit <- pmax(0, pmin(1, pit))
  pit_hist <- hist(pit, breaks = seq(0, 1, by = 0.1), plot = FALSE)
  pit_df <- data.frame(
    family = spec$family,
    midpoint = pit_hist$mids,
    density = pit_hist$density
  )

  list(
    density = density_df,
    probability = probability_df,
    quantile = quantile_df,
    random = random_df,
    pit = pit_df
  )
}

all_data <- lapply(specs, make_output_data)
density_data <- do.call(rbind, lapply(all_data, `[[`, "density"))
probability_data <- do.call(rbind, lapply(all_data, `[[`, "probability"))
quantile_data <- do.call(rbind, lapply(all_data, `[[`, "quantile"))
random_data <- do.call(rbind, lapply(all_data, `[[`, "random"))
pit_data <- do.call(rbind, lapply(all_data, `[[`, "pit"))

panel_theme <- theme(
  strip.text = element_text(size = 8),
  axis.title = element_blank(),
  panel.grid.minor = element_blank()
)
```

## Random output

Empirical CDF of random draws (`output = "random"`).

```{r random-plot, fig.width=9, fig.height=4.2}
ggplot(random_data, aes(x = x, y = y)) +
  geom_line(linewidth = 0.4, color = "#2c7fb8") +
  facet_wrap(~ family, scales = "free_x", ncol = 4) +
  labs(title = "Random output (empirical CDF thumbnails)", y = "ECDF") +
  panel_theme
```

## Probability output

CDF values (`output = "probability"`).

```{r probability-plot, fig.width=9, fig.height=4.2}
ggplot(probability_data, aes(x = x, y = y)) +
  geom_line(
    data = subset(probability_data, type == "continuous"),
    linewidth = 0.4, color = "#1b9e77"
  ) +
  geom_step(
    data = subset(probability_data, type == "discrete"),
    linewidth = 0.4, color = "#1b9e77"
  ) +
  facet_wrap(~ family, scales = "free_x", ncol = 4) +
  labs(title = "Probability output (CDF thumbnails)", y = "P(X <= x)") +
  panel_theme
```

## PIT output

PIT histogram (`output = "pit"`). For calibrated models, these are near-uniform.

```{r pit-plot, fig.width=9, fig.height=4.2}
ggplot(pit_data, aes(x = midpoint, y = density)) +
  geom_col(width = 0.085, fill = "#7570b3", alpha = 0.75) +
  geom_hline(yintercept = 1, linewidth = 0.25, linetype = "dashed") +
  facet_wrap(~ family, ncol = 4) +
  scale_x_continuous(limits = c(0, 1), breaks = c(0, 0.5, 1)) +
  labs(title = "PIT output (thumbnail histograms)", y = "Density") +
  panel_theme
```

## Density output

PDF/PMF values (`output = "density"`).

```{r density-plot, fig.width=9, fig.height=4.2}
ggplot() +
  geom_line(
    data = subset(density_data, type == "continuous"),
    aes(x = x, y = y),
    linewidth = 0.45, color = "#d95f02"
  ) +
  geom_col(
    data = subset(density_data, type == "discrete"),
    aes(x = x, y = y),
    width = 0.85, fill = "#d95f02", alpha = 0.7
  ) +
  facet_wrap(~ family, scales = "free_x", ncol = 4) +
  labs(title = "Density output (PDF/PMF thumbnails)", y = "Density / Mass") +
  panel_theme
```

## Quantile output

Inverse CDF values (`output = "quantile"`).

```{r quantile-plot, fig.width=9, fig.height=4.2}
ggplot(quantile_data, aes(x = x, y = y)) +
  geom_line(
    data = subset(quantile_data, type == "continuous"),
    linewidth = 0.4, color = "#e7298a"
  ) +
  geom_step(
    data = subset(quantile_data, type == "discrete"),
    linewidth = 0.4, color = "#e7298a"
  ) +
  facet_wrap(~ family, scales = "free_y", ncol = 4) +
  labs(title = "Quantile output (inverse-CDF thumbnails)", x = "p", y = "Q(p)") +
  panel_theme
```

## Notes

- Continuous families have `probability == pit`.
- Discrete families use randomized PIT to avoid spikes.
- The parameter values here are illustrative; the output shape changes with
  posterior draws in real model fits.
