---
title: "Running brms models with within-chain parallelization"
author: "Sebastian Weber & Paul Bürkner"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: yes
vignette: >
  %\VignetteIndexEntry{Running brms models with within-chain parallelization}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
params:
  EVAL: !r identical(Sys.getenv("NOT_CRAN"), "true")
---

```{r, SETTINGS-knitr, include=FALSE}
stopifnot(require(knitr))
options(width = 90)
opts_chunk$set(
  comment = NA,
  message = FALSE,
  warning = FALSE,
  eval = if (isTRUE(exists("params"))) params$EVAL else FALSE,
  dev = "jpeg",
  dpi = 100,
  fig.asp = 0.8,
  fig.width = 5,
  out.width = "60%",
  fig.align = "center"
)
library(ggplot2)
library(brms)
theme_set(theme_default())
```

```{r, fake-data-sim, include=FALSE, eval=TRUE}
set.seed(54647)
# number of observations
N <- 1E4
# number of group levels
G <- round(N / 10)
# number of predictors
P <- 3
# regression coefficients
beta <- rnorm(P)

# sampled covariates, group means and fake data
fake <- matrix(rnorm(N * P), ncol = P)
dimnames(fake) <- list(NULL, paste0("x", 1:P))

# fixed effect part and sampled group membership
fake <- transform(
  as.data.frame(fake),
  theta = fake %*% beta,
  g = sample.int(G, N, replace=TRUE)
)

# add random intercept by group
fake  <- merge(fake, data.frame(g = 1:G, eta = rnorm(G)), by = "g")

# linear predictor
fake  <- transform(fake, mu = theta + eta)

# sample Poisson data
fake  <- transform(fake, y = rpois(N, exp(mu)))

# shuffle order of data rows to ensure even distribution of computational effort
fake <- fake[sample.int(N, N),]

# drop not needed row names
rownames(fake) <- NULL
```

```{r, model-poisson, include=FALSE}
model_poisson <- brm(
  y ~ 1 + x1 + x2 + (1 | g),
  data = fake,
  family = poisson(),
  iter = 500, # short sampling to speedup example
  chains = 2,
  prior = prior(normal(0,1), class = b) +
    prior(constant(1), class = sd, group = g),
  backend = "cmdstanr",
  threads = threading(4),
  save_pars = save_pars(all = TRUE)
)
```

```{r, benchmark, include=FALSE}
# Benchmarks given model with cross-product of tuning parameters CPU
# cores, grainsize and iterations. Models are run with either static
# or non-static scheduler and initial values are set by default to 0 on the
# unconstrained scale. Function returns a data-frame with the
# cross-product of the tuning parameters and as result column the
# respective runtime.
benchmark_threading <- function(model, cores = 1, grainsize = 1, iter = 100,
                                static = FALSE) {

    winfo <- extract_warmup_info(model)
    sims  <- rstan::extract(model$fit)
    init <- list(extract_draw(sims, 1))

    scaling_model <- update(
        model, refresh = 0,
        threads = threading(1, grainsize = grainsize[1], static = static),
        chains = 1, iter = 2, backend = "cmdstanr"
    )

    run_benchmark <- function(cores, size, iter) {
        bench_fit <- update(
            scaling_model, warmup=0, iter = iter,
            chains = 1, seed = 1234, init = init, refresh = 0, save_warmup=TRUE,
            threads = threading(cores, grainsize = size, static = static),
            inv_metric=winfo$inv_metric[[1]],
            step_size=winfo$step_size[[1]],
            adapt_engaged=FALSE
        )
        lf <- sum(subset(nuts_params(bench_fit, inc_warmup=TRUE), Parameter=="n_leapfrog__")$Value)
        elapsed <- sum(colSums(rstan::get_elapsed_time(bench_fit$fit)))

        c(num_leapfrog=lf, runtime=elapsed)
    }

    cases <- expand.grid(cores = cores, grainsize = grainsize, iter = iter)
    res <- with(cases, mapply(run_benchmark, cores, grainsize, iter))
    cbind(cases, as.data.frame(t(res)))
}

benchmark_reference <- function(model, iter=100, init=0) {
    winfo <- extract_warmup_info(model)
    sims  <- rstan::extract(model$fit)
    init <- list(extract_draw(sims, 1))

    ref_model <- update(
        model, refresh = 0, threads = NULL,
        chains = 1, iter = 2, backend = "cmdstanr"
    )

    run_benchmark_ref <- function(iter_bench) {
        bench_fit <- update(
            ref_model, warmup=0, iter = iter_bench,
            chains = 1, seed = 1234, init = init, refresh = 0,
            inv_metric=winfo$inv_metric[[1]],
            step_size=winfo$step_size[[1]],
            adapt_engaged=FALSE
        )

        lf <- sum(subset(nuts_params(bench_fit, inc_warmup=TRUE), Parameter=="n_leapfrog__")$Value)
        elapsed <- sum(colSums(rstan::get_elapsed_time(bench_fit$fit)))

        c(num_leapfrog=lf, runtime=elapsed)
    }

    ref <- sapply(iter, run_benchmark_ref)
    ref <- cbind(as.data.frame(t(ref)), iter=iter)
    ref
}

extract_warmup_info <- function(bfit) {
    adapt  <- lapply(rstan::get_adaptation_info(bfit$fit), strsplit, split="\\n")
    step_size  <- lapply(adapt, function(a) as.numeric(strsplit(a[[1]][[1]], " = ")[[1]][2]))
    inv_metric <- lapply(adapt, function(a) as.numeric(strsplit(sub("^# ", "", a[[1]][[3]]), ", ")[[1]]))
    list(step_size=step_size, inv_metric=inv_metric)
}

extract_draw <- function(sims, draw) {
  lapply(sims, brms:::slice, dim = 1, i = draw, drop = TRUE)
}
```

## Introduction

Full Bayesian inference is a computationally very demanding task and often we
wish to run our models faster in shorter walltime. With modern computers we
nowadays have multiple processors available on a given machine such that the use
of running the inference in parallel will shorten the overall walltime. While
between-chain parallelization is straightforward by merely launching multiple
chains at the same time, the use of within-chain parallelization is more
complicated in various ways. This vignette aims to introduce the user to
within-chain parallelization with **brms**, since its efficient use depends on
various aspects specific to the users model.

## Quick summary

Assuming you have a **brms** model which you wish to evaluate faster by using
more cores per chain, for example:

```{r, eval=FALSE}
fit_serial <- brm(
  count ~ zAge + zBase * Trt + (1|patient),
  data = epilepsy, family = poisson(),
  chains = 4, cores = 4, backend = "cmdstanr"
)
```

Then running this model with threading requires `cmdstanr` as backend and you
can simply add threading support to an existing model with the `update`
mechanism as:

```{r, eval=FALSE}
fit_parallel <- update(
  fit_serial, chains = 2, cores = 2,
  backend = "cmdstanr", threads = threading(2)
)
```

The example above assumes that 4 cores are available which are best used without
within-chain parallelization by running 4 chains in parallel. When using within
chain parallelization it is still advisable to use just as many threads
*in total* as you have CPU cores. It's thus sensible in this case to reduce the
number of chains running in parallel to just 2, but allow each chain to use 2
threads. Obviously this will reduce the number of iterations in the posterior
here as we assumed a fixed amount of 4 cores.

- Only apply within-chain parallelization to large problems which take
  more than a few minutes at least to calculate. The `epilepsy`
  example above is actually too small to gain in speed (just a few seconds
  per chain on this machine).
- Within-chain parallelization is less efficient than between-chain
  parallelization. So only use within-chain parallelism if more CPUs
  can be used to run the entire analysis.
- Due to details of the model and data-set, speedups with more cores
  can be very limited. Not every model amends to within-chain
  parallelization and an empirical evaluation is in some cases
  advisable.
- Enabling threading *usually* slows down any model to some extent and
  this slowdown must be offset by sufficient cores per chain in order
  to really gain in execution speed.
- Doubling the execution speed with few cores is a lot easier than
  obtaining larger speedups with even more cores.
- Models with computationally expensive likelihoods are easier to
  parallelize than less expensive likelihoods. For example, the Poisson
  distribution involves expensive $\log\Gamma$ functions whereas the
  normal likelihood is very cheap to calculate in comparison.
- Models with many parameters (e.g., multilevel models)
  carry a large overhead when running in parallel.
- With a larger overhead of the model, the likelihood must be
  sufficiently expensive such that the relative computational cost of
  likelihood to parallelization overhead is favorable.
- Avoid using hyper-threading, that is, only use as many threads as you
  have physical cores available.
- Ensure that the data is randomly sorted such that consecutive
  subsets of the data are roughly of the same computational effort.

## Within-chain parallelization

The within-chain parallelization implemented in **brms** is based on the
`reduce_sum` facility in Stan. The basic principle that `reduce_sum` uses is to
split a large summation into arbitrary smaller partial sums. Due to the
commutativity and associativity of the sum operation these smaller partial sums
can be evaluated in any order and in parallel from one another. **brms**
leverages `reduce_sum` to evaluate the log-likelihood of the model in parallel
as for example

$$
\begin{aligned}
l(y|\theta) &= \sum_{i=1}^N l_i(y_i| \theta) \\
 &= \sum_{i=1}^{S_1} l_i(y_i| \theta) + \sum_{i=S_1+1}^N l_i(y_i| \theta).
\end{aligned}
$$

As a consequence, the within-chain parallelization requires mutually independent
log-likelihood terms which restricts its applicability to some degree.

Furthermore, the within-chain parallelization is only applicable to the
evaluation of the data likelihood while all other parts of the model, for
example priors, will remain running serially. Thus, only a partial fraction of
the entire Stan model will run in parallel which limits the potential speedup
one may obtain. The theoretical speedup for a partially in parallel running
program is described by
[Amdahl‘s law](https://en.wikipedia.org/wiki/Amdahl%27s_law).
For example, with 90% of the computational load running in parallel one can
essentially double the execution speed with 2 cores while 8 cores may only
speedup the program by at most 5x. How large the computational cost of the
log-likelihood is in relation to the entire model is very dependent on the
model of the user.

In practice, the speedups are even smaller than the theoretical speedups. This
is caused by the additional overhead implied by forming multiple smaller sums
than just one large one. For example, for each partial sum formed the entire
parameter vector $\theta$ has to be copied in memory for Stan to be able to
calculate the gradient of the log-likelihood. Hence, with more partial sums,
more copying is necessary as opposed to evaluating just one large sum. Whether
the additional copying is indeed relevant depends on the computational cost of
the log-likelihood of each term and the number of parameters. For a model with a
computationally cheap normal log-likelihood, this effect is more important than
for a model with a Poisson log-likelihood, and for multilevel models with many
parameters more copying is needed than for simpler regression models. It may
therefore be necessary to form sufficiently large partial sums to warrant an
efficient parallel execution. The size of the partial sums is referred to as the
`grainsize`, which is set to a reasonable default value. However, for some
models this tuning parameter requires some attention from the user for optimal
performance.

Finally, it is important to note that by default the exact size and order of the
partial sums is not stable as it is adjusted to the load of the system. As a
result, exact numerical reproducibility is not guaranteed by default. In order
to warrant the same size and order of the partial sums, the `static` option must
be used and set to `TRUE`, which uses a deterministic scheduler for the parallel
work.

## Example model

As a toy demonstration, we use here a multilevel Poisson model. The model is a
varying intercept model with $`r N`$ data observation which are grouped into
$`r G`$ groups. Each data item has $`r P`$ continuous covariates. The
simulation code for the fake data can be found in the appendix and it's first
$10$ rows are:

```{r}
kable(head(fake, 10), digits = 3)
```

The **brms** model fitting this data is:

```{r, eval=FALSE}
<<model-poisson>>
```

Here we have fixed the standard deviation of the between-group variation for the
intercept to the true value of $1$ as used in the simulation. This is to avoid
unfavorable geometry of the problem allowing us to concentrate on computational
aspects alone.

The Poisson likelihood is a relatively expensive likelihood due to the use of
$\log\Gamma$ function as opposed to, for example, a normal likelihood which does
is by far less expensive operations. Moreover, this example is chosen in order
to demonstrate parallelization overhead implied by a large number of parameters.

## Managing parallelization overhead

As discussed above, the key mechanism to run Stan programs with parallelization
is to split the large sum over independent log likelihood terms into arbitrary
smaller *partial sums*. Creating more *partial sums* allows to increase
simultaneous parallel computations in a granular way, but at the same time
additional overhead is introduced through the requirement to copy the entire
parameter vector for each *partial sum* formed along with further overhead due
to splitting up a single large task into multiple smaller ones.

By default, **brms** will choose a sensible `grainsize` which defines how large
a given *partial sum* will roughly be. The actual chunk size is automatically
tuned whenever the default non-static scheduler is used, which is the
recommended choice to start with. As noted before, only the static scheduler is
giving fully deterministic results since the chunk size and order of partial
sums will be the same during sampling.

While we expect that the default `grainsize` in **brms** is reasonably good for
many models, it can improve performance if one tunes the `grainsize`
specifically to a given model and data-set. We suggest to increase successively
the number of chunks a given data set is split into with the static scheduler
and run this on a single core. This way one can control the number of
*partial sum* accurately and monitor the execution time as it increases. These
experiments are run with only a single chain and very short iteration numbers as
we are not interested in the statistical results, but rather aim to be able to
explore the tuning parameter space of the chunk size as quickly as possible. The
number of iterations needed to get reliable runtime estimates for a given chunk
size will depend on many details and the easiest way to determine this is to run
this benchmark with multiple number of iterations. Whenever their results match
approximately, then the iteration numbers are sufficient. In order to
decrease the variation between runs, we also fix the random seed,
initial value and the tuning parameters of the sampler (step size and
mass matrix).

Below is an example R code demonstrating such a benchmark. The utility function
`benchmark_threading` is shown and explained in the appendix.

```{r, chunking-scale, message=FALSE, warning=FALSE, results='hide'}
chunking_bench <- transform(
    data.frame(chunks = 4^(0:3)),
    grainsize = ceiling(N / chunks)
)

iter_test <- c(10, 20, 40)  # very short test runs
scaling_chunking <- benchmark_threading(
  model_poisson,
  cores = 1,
  grainsize = chunking_bench$grainsize,  # test various grainsizes
  iter = iter_test,
  static = TRUE  # with static partitioner
)

# run as reference the model *without* reduce_sum
ref <- benchmark_reference(model_poisson, iter_test)

# for additional data munging please refer to the appendix
```

```{r, munge-chunking-scaling, include=FALSE}
scaling_chunking <- merge(scaling_chunking, chunking_bench, by = "grainsize")

single_chunk  <- transform(
    subset(scaling_chunking, chunks == 1),
    num_leapfrog_single = num_leapfrog, num_leapfrog = NULL,
    runtime_single = runtime, runtime = NULL,
    grainsize = NULL, chunks=NULL
)

scaling_chunking <- transform(
    merge(scaling_chunking, single_chunk),
    slowdown = runtime/runtime_single,
    iter = factor(iter),
    runtime_single = NULL
)

ref <- transform(ref, iter=factor(iter))
```

Graphically summarizing the results shows that with more than 8 chunks the
overhead is about 10% and increasing further with more chunks. For models
without many parameters, no such overhead should be observed. Furthermore, one
can see that 25 and 50 iterations give similar results implying that 25
iterations suffice for stable runtime estimates for these (and the following)
benchmarks. The overhead of up to 20% in this example with 16 chunks may seem
large due to the scaling of the plot. One must not forget that when we start to
use more CPU cores, the overhead is easily offset, but it limits the maximal
speedup we can get. For example, some 2 units of computation become 2.4 units
due to the overhead such that on 2 cores we don't quite double the execution
speed, but rather get a 1.6x increase in speed instead of a 2x
speedup.

Considering in addition the time per leapfrog step of the NUTS sampler
shows on an absolute scale similar information as before. The upside
of this representation is that we can visualize the slowdown in
relation to the program *without* `reduce_sum`. As we can see, the
additional overhead due to merely enabling `reduce_sum` is substantial
in this example. This is attributed in the specific example to the
large number of random effects.

```{r}
ggplot(scaling_chunking) +
    aes(chunks, slowdown, colour = iter, shape = iter) +
    geom_line() + geom_point() +
    scale_x_log10(breaks = scaling_chunking$chunks) +
    scale_y_log10(breaks=seq(0.8, 2.5, by=0.1)) +
    ggtitle("Slowdown with increasing number of chunks")

ggplot(scaling_chunking) +
    aes(chunks, 1E3 * runtime/num_leapfrog, colour = iter, shape=iter) +
    geom_line() + geom_point() +
    scale_x_log10(breaks = scaling_chunking$chunks) +
    scale_y_log10(breaks=seq(0.1, 2.0, by=0.1)) +
    geom_hline(data=ref, aes(yintercept=1E3 * runtime/num_leapfrog, colour=iter), linetype=I(2)) +
    ggtitle("Time per leapfrog step vs number of chunks",
            "Dashed line is reference model without reduce_sum") +
    ylab("Time per leapfrog step [ms]")


```

## Parallelization speedup

In practice, we are often interested in so-called "hard-scaling" properties of
the parallelization system. That is, for a fixed problem size we would like to
know how much faster we can execute the Stan program with increasing number of
threads. As nowadays CPUs usually run with so-called hyper-threading, it is also
of interest if this technique is beneficial for Stan programs as well (spoiler
alert: it's not useful). As we have seen before, the `grainsize` can have an
impact on the performance and is as such a tuning parameter. Below we
demonstrate some exemplary R code which runs a benchmark with varying number of
CPU cores and varying number of `grainsize`s.

```{r, speedup-scale, message=FALSE, warning=FALSE, results='hide'}
num_cpu <- parallel::detectCores(logical = FALSE)
num_cpu_logical <- parallel::detectCores(logical = TRUE)
grainsize_default <- ceiling(N / (2 * num_cpu))
cores <- c(2^seq(0, floor(log2(num_cpu_logical))), num_cpu, num_cpu_logical)
cores <- sort(unique(cores))
grainsize <- c(grainsize_default, grainsize_default/2, grainsize_default/4)
grainsize <- round(grainsize)

iter_scaling <- 20
scaling_cores <- benchmark_threading(
  model_poisson,
  cores = cores,
  grainsize = grainsize,
  iter = iter_scaling,
  static = FALSE
)

single_core  <- transform(
    subset(scaling_cores, cores == 1),
    runtime_single = runtime,
    num_leapfrog=NULL, runtime=NULL, cores = NULL
)

scaling_cores <- transform(
  merge(scaling_cores, single_core),
  speedup = runtime_single/runtime,
  grainsize = factor(grainsize)
)
```

It is important to consider the absolute runtime and the relative speedup vs.
running on a single core. The relative speedup can be misleading if the single
core runtime is very slow in which case speed gains on more CPUs may look overly
good. Considering instead the absolute runtime avoids this problem. After
all, we are interested in the shortest walltime we can get rather than any
relative speedups.

```{r}
ggplot(scaling_cores) +
    aes(cores, runtime, shape = grainsize, color = grainsize) +
    geom_vline(xintercept = num_cpu, linetype = 3) +
    geom_line() + geom_point() +
    scale_x_log10(breaks = scaling_cores$cores) +
    scale_y_log10(breaks=seq(0.1, 1.4, by=0.1)) +
    theme(legend.position = c(0.85, 0.8)) +
    geom_hline(data=subset(ref, iter==iter_scaling), aes(yintercept=runtime), linetype=I(2)) +
    ggtitle("Runtime with varying number of cores",
            "Dashed line is reference model without reduce_sum")

ggplot(scaling_cores) +
  aes(cores, speedup, shape = grainsize, color = grainsize) +
  geom_abline(slope = 1, intercept = 0, linetype = 2) +
  geom_vline(xintercept = num_cpu, linetype = 3) +
  geom_line() + geom_point() +
  scale_x_log10(breaks=scaling_cores$cores) +
  scale_y_log10(breaks=scaling_cores$cores) +
  theme(aspect.ratio = 1) +
  coord_fixed(xlim = c(1, num_cpu_logical), ylim = c(1, num_cpu_logical)) +
  ggtitle("Relative speedup vs 1 core")
```

The vertical dotted line marks the physical number of CPU cores on the machine
this was run. The horizontal dashed line in the plot with absolute
runtime marks the respective runtime of the model *without*
`reduce_sum` and the dashed unity line in the plot with the relative
speedup marks the theoretical maximal speedup. We can see
that there is no further reduction in execution time when increasing the thread
count to be greater than the number of physical CPUs. Hence, the use of
hyper-threading is not helpful when aiming to maximize the speed of a Stan
program. Moreover, the use of threading outperforms the single core
runtime only when using more than 4 cores in this example.

For this example, the shown `grainsize`s matter on some machines but
not on others, so your results may look quite different from what is shown here.
The overall speedups may not seem impressive in this case, which is attributed
in this case to the large number of parameters relative to the number of
observations. However, we can still outperform the single core
runtime when using many cores. Though the most important advantage of
threading is that with an increasing data set size, the user has the option to
use a brute-force approach to balance the increase in walltime needed.

```{r}
kable(scaling_cores, digits = 2)
```

For a given Stan model one should usually choose the number of chains and the
number of threads per chain to be equal to the number of (physical) cores one
wishes to use. Only if different chains of the model have relatively different
execution times (which they should not have, but it occurs sometimes in
practice), then one may consider the use of hyper-threading. Doing so will share
the resources evenly across all chains and whenever the fastest chain finishes,
the freed resources can be given to the still running chains.

## Appendix

### Fake data simulation

```{r, eval=FALSE}
<<fake-data-sim>>
```

### Poisson example model

```{r, eval=FALSE}
<<model-poisson>>
```

### Threading benchmark function

```{r, eval=FALSE}
<<benchmark>>
```

### Munging of slowdown with chunking data

```{r, eval=FALSE}
<<munge-chunking-scaling>>
```
