---
title: "Estimating Phylogenetic Multilevel Models with brms"
author: "Paul Bürkner"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
---

<!--
%\VignetteEngine{knitr::rmarkdown}
%\VignetteIndexEntry{Fit phylogenetic models with brms}
%\VignetteEncoding{UTF-8}
-->

```{r, SETTINGS-knitr, include=FALSE}
stopifnot(require(knitr))
opts_chunk$set(
  eval = FALSE,
  fig.align='center', fig.width = 7, fig.height = 3
)
```

## Introduction

In the present vignette, we want to discuss how to specify phylogenetic multilevel models using **brms**. These models are relevant in evolutionary biology when data of many species are analyzed at the same time. The usual approach would be to model species as a grouping factor in a multilevel model and estimate varying intercepts (and possibly also varying slopes) over species. However, species are not independent as they come from the same phylogenetic tree and we thus have to adjust our model to incoporate this dependence. The examples discussed here are from the book *Modern Phylogenetic Comparative Methods and the application in Evolutionary Biology* (Garamszegi, 2014) more specifically from the online practice material of Chapter 11. The necessary data can be downloaded from  http://www.mpcm-evolution.org/practice/online-practical-material-chapter-11. Some of these models may take a few minutes to fit.


## A Simple Phylogenetic Model

Assume we have measurements of a phenotype `phen` (say the body size) and a `cofactor` variable (say the temperature of the environment). We prepare the data using the following code.


```{r}
setwd("<insert path to data folder here>")
library(ape)
library(MCMCglmm)
phylo <- ape::read.nexus("phylo.nex")
data_simple <- read.table("data_simple.txt", header = TRUE)
head(data_simple)
```

The packages **ape** and **MCMCglmm** are required only for the data preparation and not for actually fitting the models. The `phylo` object contains information on the relationship between species and we have to bring it into the right format that is a covariance matrix of species.

```{r}
inv.phylo <- MCMCglmm::inverseA(phylo, nodes = "TIPS", scale = TRUE)
A <- solve(inv.phylo$Ainv)
rownames(A) <- rownames(inv.phylo$Ainv)
```

In contrast to **MCMCglmm**, **brms** requires the covariance matrix not its inverse. Now, we are ready to fit our first phylogenetic multilevel model:

```{r}
model_simple <- brm(phen ~ cofactor + (1|phylo), data = data_simple, 
                    family = gaussian(), cov_ranef = list(phylo = A),
                    prior = c(prior(normal(0, 10), "b"),
                              prior(normal(0, 50), "Intercept"),
                              prior(student_t(3, 0, 20), "sd"),
                              prior(student_t(3, 0, 20), "sigma")))
```

With the exception of `cov_ranef = list(phylo = A)` this is a basic multilevel model with a varying intercept over species (`phylo` is an indicator of species in this data set). However, by using the `cov_ranef` argument, we make sure that species are correlated as specified by the covariance matrix `A`. Setting priors is not required for achieving good convergence for this model, but it improves sampling speed a bit. After fitting, the results can be investigated using the among others the following methods.

```{r}
summary(model_simple)
plot(model_simple)
plot(marginal_effects(model_simple), points = TRUE) 
```

The so called phylogenetic signal (often symbolize by $\lambda$) can be computed as follows and will be roughly $\lambda = 0.7$ for this example.

```{r}
hyp <- "sd_phylo__Intercept^2 / (sd_phylo__Intercept^2 + sigma^2) = 0"
(hyp <- hypothesis(model_simple, hyp, class = NULL))
plot(hyp)
```


## A Phylogenetic Model with Repeated Mesurements

Often, we have multiple observations per species and this allows to fit more complicated phylogenetic models.

```{r}
data_repeat <- read.table("data_repeat.txt", header = TRUE)
data_repeat$spec_mean_cf <- 
  with(data_repeat, sapply(split(cofactor, phylo), mean)[phylo])
head(data_repeat)
```

The variable `spec_mean_cf` just contains the mean of the cofactor of each species. The code for the repeated measurement phylogenetic model looks as follows:

```{r}
model_repeat1 <- brm(phen ~ spec_mean_cf + (1|phylo) + (1|species), 
                     data = data_repeat, family = gaussian(), 
                     cov_ranef = list(phylo = A),
                     prior = c(prior(normal(0,10), "b"),
                               prior(normal(0,50), "Intercept"),
                               prior(student_t(3,0,20), "sd"),
                               prior(student_t(3,0,20), "sigma")),
                     sample_prior = TRUE, chains = 2, cores = 2, 
                     iter = 4000, warmup = 1000)
```

The variables `phylo` and `species` are identical as they are both identifiers of the species. However, we model the phylogenetic covariance only for `phylo` and thus the `species` variable accounts for any specific effect that would be independent from the phylogenetic relationship between species (e.g., environmental or niche effects). Again we can obtain model summaries as well as estimates of the phylogenetic signal.

```{r}
summary(model_repeat1)
plot(model_repeat1)
plot(marginal_effects(model_repeat1), points = TRUE)
```

```{r}
hyp <- paste(
  "sd_phylo__Intercept^2 /", 
  "(sd_phylo__Intercept^2 + sd_species__Intercept^2 + sigma^2) = 0"
)
(hyp <- hypothesis(model_repeat1, hyp, class = NULL))
plot(hyp, chars = NULL)
```

So far, we have completely ignored the variability of the cofactor within species. To incorporate this into the model, we define

```{r}
data_repeat$within_spec_cf <- data_repeat$cofactor - data_repeat$spec_mean_cf
```

and then fit it again using `within_spec_cf` as an additional predictor.

```{r}
model_repeat2 <- update(model_repeat1, formula = ~ . + within_spec_cf,
                        newdata = data_repeat, chains = 2, cores = 2, 
                        iter = 4000, warmup = 1000)
```

The results are almost unchanged, with apparently no relationship between the phenotype and the within species variance of `cofactor`.

```{r}
summary(model_repeat2)
plot(model_repeat2, N = 6)
plot(marginal_effects(model_repeat2), points = TRUE) 
```

Also, the phylogenetic signal remains more or less the same.

```{r}
hyp <- paste(
  "sd_phylo__Intercept^2 /", 
  "(sd_phylo__Intercept^2 + sd_species__Intercept^2 + sigma^2) = 0"
)
(hyp <- hypothesis(model_repeat2, hyp, class = NULL))
plot(hyp, chars = NULL)
```


## A Phylogenetic Meta-Analysis

Let's say we have Fisher's z-transformated correlation coefficients $Zr$ per species along with corresponding sample sizes (e.g., correlations between male coloration and reproductive success): 

```{r}
data_fisher <- read.table("data_effect.txt", header = TRUE)
data_fisher$obs <- 1:nrow(data_fisher)
head(data_fisher)
```

We assume the sampling variance to be known and as $V(Zr) = \frac{1}{N - 3}$ for Fisher's values, where $N$ is the number of individuals per species. Incorporating the known sampling variance into the model is straight forward. One has to keep in mind though, that **brms** requires the sampling standard deviation (square root of the variance) as input instead of the variance itself. The group-level effect of `obs` represents the residual variance, which we have to model explicitly in a meta-analytic model.

```{r}
model_fisher <- brm(Zr | se(sqrt(1 / (N - 3))) ~ 1 + (1|phylo) + (1|obs), 
                    data = data_fisher, family = gaussian(), 
                    cov_ranef = list(phylo = A),
                    prior = c(prior(normal(0, 10), "Intercept"),
                              prior(student_t(3, 0, 10), "sd")),
                    control = list(adapt_delta = 0.95),
                    chains = 2, cores = 2, iter = 4000, warmup = 1000)
```

A summary of the fitted model is obtained via

```{r}
summary(model_fisher)
plot(model_fisher)
```

The meta-analytic mean (i.e., the model intercept) is $0.16$ with a credible interval of $[0.07, 0.24]$. Thus the mean correlation across species is positive according to the model.


## A phylogenetic count-data model

Suppose that we analyze a phenotype that consists of counts instead of being continuous. In such a case, the assumption of a normal distribution will likely not be justified and it is recommended to use a distribution explicitely suited for count data, such as the Poisson distribution. The following data set provides an example.

```{r}
data_pois <- read.table("data_pois.txt", header = TRUE)
data_pois$obs <- 1:nrow(data_pois)
head(data_pois)
```

Here, we fit the same simple phylogenetic model as in the first example, but specify `family = poisson()`.

```{r}
model_pois <- brm(phen_pois ~ cofactor + (1|phylo) + (1|obs), 
                  data = data_pois, family = poisson("log"), 
                  cov_ranef = list(phylo = A),
                  prior = c(set_prior("normal(0,5)"),
                            set_prior("normal(0,5)", "Intercept"),
                            set_prior("student_t(3,0,5)", "sd")),
                  chains = 2, cores = 2, iter = 4000, warmup = 1000)
```

Again, we obtain a summary of the fitted model via

```{r}
summary(model_pois)
plot(model_pois)
plot(marginal_effects(model_pois), points = TRUE) 
```

Similar to the first example, we see that `cofactor` has a positive relationship with the phenotype. One should note, however, that the estimates of the poisson model are on the log-scale as we applied the canonical log-link function in this example.
